### 2017-10-21 paper/paper.tex
Start adding paper.

\documentclass[sigchi]{acmart}

\usepackage{booktabs}
\usepackage{commath}
\usepackage{threeparttable}

% --- Symbols
\newcommand\symUserSet{U}
\newcommand\symItemSet{I}
\newcommand\symUserFeaturesSet{F^U}
\newcommand\symItemFeaturesSet{F^I}
\newcommand\symItemEmbeddingMatrix{E^I}
\newcommand\symUserEmbeddingMatrix{E^U}

\newcommand\symUserInteractionSet{S}

\DeclareMathOperator{\sign}{sign}
\renewcommand\vec{\mathbf}

\begin{document}

\title{Parameter-efficient multi-modal embeddings for representing diverse user interests}
\author{Maciej Kula}
\email{maciej.kula@gmail.com}
\date{\today}
\settopmatter{printacmref=false, printccs=false, printfolios=true}
\setcopyright{none}
\acmConference[]{}{}{}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10002951.10003260.10003261.10003269</concept_id>
<concept_desc>Information systems~Collaborative filtering</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10002951.10003317.10003338.10003346</concept_id>
<concept_desc>Information systems~Top-k retrieval in databases</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Information systems~Collaborative filtering}
\ccsdesc[500]{Information systems~Top-k retrieval in databases}

\keywords{Recommender Systems, Matrix Factorization, Binary Representations}

\begin{abstract}
Most existing recommendation approaches implicitly treat user tastes as unimodal, resulting in an average-of-tastes representations when multiple distinct interests are present. We show that appropriately modelling the multi-faceted nature of user tastes through a mixture-of-tastes model leads to large increases in recommendation quality at a very modest cost in model complexity. Our result holds both for deep sequence-based and traditional factorization models. 
\end{abstract}

\maketitle

\section{Introduction}

\section{Experiments}
\subsection{Datasets}
For our experiments, we use the following datasets.

\paragraph{Movielens 10M} A dataset of 10 million ratings across 10,000 movies and 72,000 users \citep{harper2016movielens}.
\paragraph{Goodbooks} A dataset of 6 million ratings across 53,000 users and 10,000 most popular books from the Goodbooks online book recommendation and sharing service \citep{goodbooks2017}.
\paragraph{Amazon} A dataset of ratings and reviews gathered from the Amazon online shopping service. After pruning users and items with fewer than 10 ratings, the dataset contains approximately 4 million ratings from 100,000 users over 114,000 items.

We treat all datasets as implicit feedback datasets, where the existence of an edge between a user and an item expresses implicit preference, and the lack of an edge implicit lack of preference.

\subsection{Experimental setup}
For factorization models, we split the interaction datasets randomly into train, validation, and test sets. We use 80\% of interactions for training, and 10\% each for validation and testing. We make no effort to ensure that all items and users in the validation and test sets have a minimum number of training interactions. Our results therefore represent partial cold-start conditions.

For sequence-based models, we order all interactions chronologically, and split the dataset by randomly assigning users into train, validation, and test sets. This means that the train, test, and validation sets are disjoint along the user dimension. For each dataset, we define a maximum interaction sequence length. This is set to 100 for the Goodbooks and Movielens datasets, and 50 for the Amazon dataset, as the interaction sequences in the Amazon dataset are generally shorter. Sequences shorter than the maximum sequence length are padded with zeros.

We use mean reciprocal rank (MRR) as our measure of model quality. In factorization models, we use the user representations obtained from the training set to construct rankings over items in the test set. In sequence models, we use the last element of the test interaction sequence as the prediction target; the remaining elements are used to compute the user representation.

We perform extensive hyperparameter optimization across both our proposed models and all baselines to mitigate the danger that our results driven either by models' sensitivity to hyperparameter choices (when a superior model performs poorly because of a poor choice of hyperparameters), or experimenter bias (in favouring the optimization of the novel model). We optimize batch size, number of training epochs, the learning rate, L2 regularization weight, the loss function, and (where appropriate) the number of taste mixture components.

We experiment with two loss functions:
\begin{itemize}
\item Bayesian personalised ranking (BPR, \citet{rendle2009bpr}), and
\item adaptive sampling maximum margin loss, following \citet{weston2011wsabie}.
\end{itemize}
For both loss functions, for any known positive user-item interaction pair $(u, i)$, we uniformly sample an implicit negative item $j \in \symUserInteractionSet^-$. For BPR, the loss for any such triplet is given by
\begin{equation}
1 - \sigma\left(r_{ui} - r_{uj}\right),
\end{equation}
where $\sigma$ denotes the sigmoid function.
The adaptive sampling loss is given by
\begin{equation}
\abs{1 - r_{ui} + r_{uj}}_{+}.
\end{equation}
For any $(u, i)$ pair, if the sampled negative item $j$ results in a zero loss (that is, the desired pairwise ordering is not violated), a new negative item is sampled, up to a total of $k$ attempts. This leads the model to perform more gradient updates in areas where its ranking performance is poorest.

Across all of our experiments, the adaptive maximum margin loss consistently outperforms the BPR loss on both baseline and mixture models. We therefore only report results for the adaptive loss.

\begin{table*}[htbp]
\begin{threeparttable}
\caption{Experimental results}
\label{tb:results}
\begin{tabular}{lrrr}
\toprule
 Model        &   Movielens 10M &   Amazon &   Goodbooks-10K \\
\midrule
 LSTM         &            0.09 &    0.15  &           0.116 \\
 Mixture-LSTM &            0.1  &    0.189 &           0.136 \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small{
\item[1] Ratio of binary model MRR to real-valued model MRR
\item[2] Predictions Per Millisecond: how many items can be scored per millisecond
\item[3] Ratio of binary PPMS to real-valued PPMS
\item[4] Ratio of memory required to store binary vs. real-valued parameters
}
\end{tablenotes}
\end{threeparttable}
\end{table*}

%% \section{Introduction}
%% Industry ranking and recommendation systems need to scale to tens or hundreds of millions of items and users. As the number of items available to be recommended grows large, efficiently ranking the item catalogue to produce top-ranked recommendations becomes increasingly challenging. With large catalogues, ranking latency (in on-line settings) and compute and storage costs (in pre-compute settings) place significant constraints on the design of the entire recommender system.

%% In an on-line setting, where recommendation latency directly impacts the user experience, system designers have to work within a strict latency budget. Respecting that budget often means trading off model accuracy for speed, either via smaller and less expressive models (with lower-dimensionality representations) or through aggressive use of heuristics to exclude large classes of candidate items from scoring. For example, Pinterest uses a combination of heuristics and candidate-generation models to select approximately 1000 pins, from billions of available pins, as input to its related pins ranking system \citep{liu2017related}. Similarly, approximate nearest neighbour search techniques are used at YouTube \citep{covington2016deep} to reduce the number of candidate videos that need to be scored for each user.

%% The challenge of low-latency scoring leads many systems to rely on a pre-compute system, where recommendations are calculated in advance and stored for serving at a later time. The primary disadvantage of such a system is its inability to react dynamically to user actions (such as new interactions that allow the recommendations to be refined) and recommendation context (location, time of day or year). This leads to often complex architectures that mix fairly simple on-line models, computationally expensive offline models, and intermediate complexity near-line algorithms where model updating is important \citep{amatriain2013big}.

%% Additionally, while a pre-compute solution removes model prediction from the critical path, it still requires substantial investment in computational and storage resources to calculate and cache the predictions of the model. Often the amount of computation required makes timely updating of the cached results impossible. At Pinterest, pre-calculation of related pins results was so demanding that ranking of different segments of the catalogue had to be staggered in time, leading to stale results and reduced ease of development \citep[Section 5.5]{liu2017related}. \citet{koenigstein2012efficient} estimate that computing recommendations for the Yahoo! Music dataset \citep{dror2011yahoo} with a 50-dimensional latent factor model woud take over 135 hours.

%% To alleviate these constraints in both online and offline settings, we evaluate the use of binary item and user representations in learning-to-rank matrix factorization models. Following the approach of \citet{rastegari2016xnor}, we estimate binary user and item representations that are several times faster to score and requiring a fraction of the memory (for the same latent dimensionality). Unfortunately, they are substantially less accurate than standard learning-to-rank approaches. The resulting speed-accuracy trade-offs favour simply reducing the latent dimensionality of traditional models when evaluation speed is desired.

%% We speculate that binary representations are more useful in case of more complex models (such as deep convolutional models) where a single prediction requires many more floating operations than relatively compact bilinear recommender models.

%% \section{Binary latent representations}
%% \label{sec:approach}
%% \citet{rastegari2016xnor} introduce XNOR-Networks, an approach that combines 1-bit quantization of fully-connected and convolutional layers with representing dot products through scaled XNOR and bitcounting operations.

%% We adapt their approach to the recommendation task by learning binary versions of user and item vectors within a learning-to-rank factorization model. This allows us to use binary instead of floating point operations for computing recommendation rankings, which has significant speed and memory use advantages. In a real-valued $n$-dimensional factorization model, item and user representations take $4n$ bytes of memory, and computing a single prediction takes $2n$ floating point operations. In an equivalent binary model, $\frac{3}{32}n$ binary operations are required for each prediction (XOR, negation, and bitcounting), and the representations take $\frac{1}{8}n$ bytes of memory. Naive calculations suggest that, for the same latent dimensionality, a binary model would be over 20 times faster to score, and take less then 5\% of memory to store its representations.

%% Naturally, using binary instead of real-valued representations entails a loss of representational power, and the resulting models can be expected to be less accurate. Nevertheless, binary models may still be very useful in a production system, for two reasons. Firstly, they may offer the system designer a more favourable trade-off between speed and accuracy than simply changing the number of latent dimensions in a real-valued model. If latency is the binding constraint, switching to a binary model may offer better or comparable speed with better ranking performance. If accuracy is the priority, a binary model may offer equivalent ranking performance while being faster to score. Secondly, many production systems resort to heavy use of heuristics when selecting candidates to be scored by their ranking systems (as scoring all candidates would be infeasible). To the extent that these heuristics are suboptimal, replacing them with a binary representation model would improve the overall accuracy of the system.

%% To quantify the trade-offs involved, we construct and fit both a real-valued and a binary version of simple implicit learning-to-rank factorization model. We introduce the notation and the models below.

%% Let $\symUserSet$ be the set of users, and $\symItemSet$ be the set of items. Each user interacts with a number of items $\symUserInteractionSet^+$; the set of all remaining items is denoted by $\symUserInteractionSet^-$.

%% In a traditional real-valued latent factor model, the prediction $r_{ui}$ for any user-item interaction pair $(u, i) \in \symUserSet \times \symItemSet$ is given by
%% \begin{equation}
%% r_{ui} = \vec{u}_u \cdot \vec{i}_i + b_u + b_i
%% \end{equation}
%% where $\vec{u}_u$ denotes the user and $\vec{i}_i$ the item $n$-dimensional latent vectors, and $b_u$ and $b_i$ the user and item biases.

%% \citet{rastegari2016xnor} show that the dot product of two real-valued vectors can be approximated in the binary domain by (1) binarizing the input vectors using the sign function (so that the results lie in $\{1, -1\}$), (2) taking their dot product, and (3) scaling the result by the average magnitude of the vectors' elements. Letting
%% \begin{equation}
%% \alpha = \frac{1}{n}\norm{\vec{i}_{i}}_{\ell1}
%% \end{equation}
%% and
%% \begin{equation}
%% \beta = \frac{1}{n}\norm{\vec{u}_{u}}_{\ell1}
%% \end{equation}
%% represent the scaling factors, the approximation is given by
%% \newcommand\binaryApproximation{ \alpha \beta\left(\sign(\vec{u}_u) \cdot \sign(\vec{i}_i)\right)}
%% \begin{equation}
%% \vec{u}_u \cdot \vec{i}_i \approx \binaryApproximation
%% \end{equation}
%% Applying this to the latent-factor model, the prediction for a user-item pair is given by
%% \begin{equation}
%% r_{ui} = \binaryApproximation + b_u + b_i
%% \end{equation}
%% Note that the scaling factors $\alpha$ and $\beta$ need to be stored (in addition to the latent representations and biases) to compute predictions in the binary model. Their use also implies a constant cost of two floating-point multiplications per dot product when using binary representations.

%% The model is trained using backpropagation, and so we use a smooth approximation to the derivative of the sign function to facilitate training. Following \citet{courbariaux2016binarized}, we define
%% \begin{equation}
%%   \frac{d\sign}{dw}=
%%   \begin{cases}
%%     1 & \text{if}\ \abs{w} \leq 1 \\
%%     0 & \text{otherwise}.
%%   \end{cases}
%% \end{equation}

%% We use two loss functions to train the model:
%% \begin{itemize}
%% \item Bayesian personalised ranking (BPR, \citet{rendle2009bpr}), and
%% \item adaptive sampling maximum margin loss, following \citet{weston2011wsabie}.
%% \end{itemize}
%% For both loss functions, for any known positive user-item interaction pair $(u, i)$, we uniformly sample an implicit negative item $j \in \symUserInteractionSet^-$. For BPR, the loss for any such triplet is given by
%% \begin{equation}
%% 1 - \sigma\left(r_{ui} - r_{uj}\right),
%% \end{equation}
%% where $\sigma$ denotes the sigmoid function.
%% The adaptive sampling loss is given by
%% \begin{equation}
%% \abs{1 - r_{ui} + r_{uj}}_{+}.
%% \end{equation}
%% For any $(u, i)$ pair, if the sampled negative item $j$ results in a zero loss (that is, the desired pairwise ordering is not violated), a new negative item is sampled, up to a total of $k$ attempts. This leads the model to perform more gradient updates in areas where its ranking performance is poorest.

%% For the purposes of this paper, we focus on a simple bilinear collaborative filtering model, and do not use any external metadata information or model the sequential nature of the data. However, the binary dot product approach generalizes beyond simple factorization models, and can be applied as a component of any model whose final scoring step involves a dot product between user and item representations. In particular, models using recurrent \citep{hidasi2015session} or convolutional \citep{lynch2015images} item or user representations can easily be augmented to use binary dot products in the final ranking stages.

%% \section{Experiments}
%% \begin{table*}[htbp]
%% \begin{threeparttable}
%% \caption{Movielens 1M results}
%% \label{tb:results}
%% \centering
%% \begin{tabular}{rrrrrrrr}
%% \toprule
%% Dimension &   MRR & Binary MRR & MRR ratio\tnote{1} &   PPMS\tnote{2} & Binary PPMS & PPMS ratio\tnote{3} & Memory use ratio\tnote{4} \\
%% \midrule
%%        32 & 0.077 &      0.059 &     0.768 & 87,758 &     264,146 &      3.010 &            0.091 \\
%%        64 & 0.075 &      0.066 &     0.878 & 45,992 &     220,833 &      4.802 &            0.062 \\
%%       128 & 0.076 &      0.071 &     0.935 & 23,870 &     166,252 &      6.965 &            0.047 \\
%%       256 & 0.078 &      0.071 &     0.908 & 12,284 &     190,131 &     15.477 &            0.039 \\
%%       512 & 0.080 &      0.073 &     0.912 &   6074 &     122,637 &     20.190 &            0.035 \\
%%      1024 & 0.080 &      0.075 &     0.936 &   3056 &      61,301 &     20.056 &            0.033 \\
%% \bottomrule
%% \end{tabular}
%% \begin{tablenotes}
%% \small{
%% \item[1] Ratio of binary model MRR to real-valued model MRR
%% \item[2] Predictions Per Millisecond: how many items can be scored per millisecond
%% \item[3] Ratio of binary PPMS to real-valued PPMS
%% \item[4] Ratio of memory required to store binary vs. real-valued parameters
%% }
%% \end{tablenotes}
%% \end{threeparttable}
%% \end{table*}
%% To assess the accuracy-speed trade-offs enabled by the XNOR approach, we conduct an experiment on the Movielens 1M dataset \citep{harper2016movielens}. The dataset contains 1 million ratings from 6000 users on 4000 movies. Because the computational speed improvements offered by the binary representations are linear in the size of the catalogue, the results on this dataset should be indicative of results that can be obtained on larger datasets. At the same time, the dataset's small size enables us to easily test hyperparameter configurations and estimate models with high latent dimensions.

%% \subsection{Experimental setup}
%% We randomly divide the dataset into a training, test, and validation set. In order to build a picture of the accuracy-speed trade-offs afforded by real-valued and binary latent models, we explicitly build models for between 32 and 1024 latent dimensions. For every latent dimensionality, we conduct a random search over the hyperparameter space, and pick optimal initial learning rate, loss function, L2 penalty, minibatch size and number of training epochs for each algorithm based on their performance on the test set. The final results are obtained from ranking interactions from the validation set. We use mean reciprocal rank (MRR) as a measure of ranking quality.

%% Benchmark results are measured on an Intel Xeon E5-2686v4 CPU by running a 500 repetitions of scoring 100,000 items and averaging the results.

%% \subsection{Implementation}
%% The model is implemented in PyTorch and trained using the nVidia K40 GPUs. The implementation is only marginally more complex than a standard learning-to-rank model, and is accomplished by simply replacing the user-item vector dot product operation with its binary counterpart. The change results in a small increase in training time.

%% During training, the embedding parameters are stored and updated as single precision floating point values. Similarly, the XNOR dot product is carried out using floats in $\{1, -1\}$. The initial learning rate, loss function, L2 penalty, minibatch size and number of training epochs are treated as model hyperparameters. All models are trained using the Adam training rate schedule \citep{kingma2014adam}.

%% The prediction code runs on the CPU and is implemented in C using Intel X86 AVX2 SIMD intrinsics. SIMD (Single Instruction Multiple Data) instructions allow the CPU to operate on multiple pieces of data in parallel, achieving significant speedups over the scalar version. We use explicit intrinsics rather than compiler autovectorization to ensure that neither the real-valued nor the binary prediction code is unfairly disadvantaged by the quality of compiler autovectorization.

%% The real-valued prediction code is implemented using 8-float wide fused multiply-add instructions (\texttt{\_mm256\_fmadd\_ps}). In the binary version, the real-valued embedding parameters used in training are discarded, and the derived 1-bit weights are packed into 32-bit integer buffers. The XNOR dot product is implemented using 8-integer wide XOR operations (256 binary weights are processed at a time), followed by a popcount instruction to count the number of on bits in the result. We use the \texttt{libpopcnt} \citep{mula2016faster} library for the bit counting operations.

%% Both versions use 32-bit aligned input data to utilise aligned SIMD register load instructions. The code is compiled using GCC 4.8.4 for the AVX2-enabled Broadwell architecture.

%% The source code for both model training and prediction is available on Github at \url{https://github.com/maciejkula/binge}.

%% \subsection{Results}
%% \label{sec:results}
%% Table \ref{tb:results} summarises the results of the Movielens 1M experiment. Each row presents the best MRR (mean reciprocal rank) results for both real-valued and binary models of a given dimensionality. MRR ratio denotes the fraction of the real-valued MRR the binary model achieves for that dimensionality; PPMS (predictions per millisecond) ratio denotes how many more predictions the binary model can compute per millisecond.

%% Results on scoring speed and memory use broadly confirm the naive calculations from section \ref{sec:approach}, converging to over 20 times faster scoring at around 3\% of memory as latent dimensionality increases. Memory savings plateau as latent dimensionality increases and storing the binary scaling factors becomes less important.

%% As expected, for the same dimensionality a binary model achieves lower ranking accuracy. On average, the accuracy loss when moving from a continuous to a binary model of the same latent dimensionality is around 11\%, varying between 6\% and 23\%.

%% Unfortunately, while continuous models retain good accuracy as latent dimensionality decreases, binary models' representational power sharply deteriorates. Moving from the 1024 to 32 dimensions in the continuous model implies a 29 times increase in prediction speed at the expense of a modest 4\% decrease in accuracy. This compares favourably to switching to binary representations: moving to a 1024-binary dimensional representation implies a sharper accuracy drop at 6\% in exchange for a smaller 20 times increase in prediction speed. Moving to 32 dimensions yields a further speed gains at 86 times, but at the cost of a considerable loss of accuracy at 26\%.

%% The results suggest that latent recommender models have good representational power even at low latent dimensionalities. This advantage is lost when using binary embeddings, which need to be high-dimensional to achieve comparable accuracy. At the same time, binary representations' speed advantage is only evident at high dimensions: at low dimensions, only floating-point operations can use enjoy the advantages of SIMD operations, and the fixed overhead of binary scaling factors $\alpha$ and $\beta$ constitutes a larger proportion of the total computational cost. We conjecture that the attractiveness of binary representations is tightly coupled with high-dimensional models where a single prediction requires many floating point operations, such as convolutional neural networks. Relatively compact latent factor models do not fall into this category.

%% \section{Alternative approaches}
%% Given the limited success of binary embedding, a more promising approach is offered by existing work that focuses on reducing the number of dot products which need to be performed in order to retrieve top recommendations. \citet{koenigstein2012efficient} introduces an exact branch-and-bound based on metric trees as well as an approximate algorithm that clusters users and serves recommendations computed for the cluster centers. \citet{shrivastava2014asymmetric} extend well-known \citep{indyk1998approximate} locality-sensitive hashing techniques to maximum inner product search (MIPS) through asymmetric transformations of the query and candidate vectors.

%% The method that is most closely related to the model we present in this paper is \citet{Shen_2015_ICCV}, who expand on MIPS by learning asymmetric binary hash functions. The hash functions are trained to minimize the L2 norm between the inner product matrix of the original input vectors and the inner product matrix of their binary representations.

%% \section{Conclusion}
%% While prediction latency is a pressing concern for many recommender systems, we find that the already compact latent factor models do not stand to benefit from using binary latent representations. The sharp drop in representational power exhibited by binary embeddings makes other approaches (such as exact \citep{koenigstein2012efficient} and approximate \citep{shrivastava2014asymmetric} maximum inner product search) more promising avenues when optimizing large-scale ranking.

\bibliography{bibliography}
\bibliographystyle{ACM-Reference-Format}

\end{document}

### 2017-10-21 experiment/summarize_results.py
Add code for generating hyperparam senitivity table.

@@ -1,0 +1,11 @@
+from mixture.results import (generate_hyperparameter_table,
+                             generate_performance_table,
+                             read_results)
+
+
+if __name__ == '__main__':
+
+    results = read_results('mixture_results', 'sequence')
+
+    print(generate_performance_table(results))
+    print(generate_hyperparameter_table(results))

### 2017-10-21 mixture/results.py

@@ -31,9 +31,13 @@ def summarize_trials(trials):
             print('Best test {}: {}'.format(model_type, results[0]['result']))
 
 
-def _get_best_test_result(results, model_type):
+def _get_best_test_result(results, model_type, filter_fnc=None):
 
-    results = [x for x in results.trials if _get_type(x) == model_type]
+    if filter_fnc is None:
+        filter_fnc = lambda x: True
+
+    results = [x for x in results.trials
+               if _get_type(x) == model_type and filter_fnc(x)]
     best = sorted(results, key=lambda x: x['result']['validation_mrr'])[-1]
 
     return best['result']['test_mrr']
@@ -60,12 +64,36 @@ def generate_performance_table(results):
     headers = ['Model', 'Movielens 10M', 'Amazon', 'Goodbooks-10K']
     rows = []
 
-    for (model_name, model) in (('LSTM', 'lstm'), ('Mixture-LSTM', 'mixture')):
+    for (model_name, model) in (('LSTM', 'lstm'),
+                                ('Mixture-LSTM', 'mixture')):
 
         row = [model_name]
 
         for dataset in ('10M', 'amazon', 'goodbooks'):
             mrr = _get_best_test_result(results[dataset], model)
+            row.append(mrr)
+
+        rows.append(row)
+
+    return tabulate(rows,
+                    headers=headers,
+                    floatfmt='.4f',
+                    tablefmt='latex_booktabs')
+
+
+def generate_hyperparameter_table(results):
+
+    headers = ['Mixture components', 'Movielens 10M', 'Amazon', 'Goodbooks-10K']
+    rows = []
+
+    for num_components in (2, 4, 6, 8):
+
+        row = [num_components]
+
+        filter_fnc = lambda x: x['result']['hyper'].get('num_components') == num_components
+        
+        for dataset in ('10M', 'amazon', 'goodbooks'):
+            mrr = _get_best_test_result(results[dataset], 'mixture', filter_fnc)
             row.append(mrr)
 
         rows.append(row)

### 2017-10-22 mixture/factorization_representation.py
Switch to built-in softmax.
For numerical stability.

@@ -1,16 +1,11 @@
+import random
+
 import torch
 
 import torch.nn as nn
+import torch.nn.functional as F
 
 from spotlight.layers import ScaledEmbedding, ZeroEmbedding
-
-
-def softmax(x, dim):
-
-    numerator = torch.exp(x)
-    denominator = numerator.sum(dim, keepdim=True).expand_as(numerator)
-
-    return numerator / denominator
 
 
 class MixtureNet(nn.Module):
@@ -94,12 +89,22 @@ class MixtureNet(nn.Module):
                                   self.num_components,
                                   embedding_size))
         item_embedding = item_embedding.unsqueeze(1).expand_as(user_attention)
-        attention = (softmax((user_attention * item_embedding).sum(2), 1))
+
+        attention = F.softmax((user_attention * item_embedding).sum(2))
+
         preference = ((user_tastes * item_embedding)
                       .sum(2))
         weighted_preference = (attention * preference).sum(1).squeeze()
 
         user_bias = self.user_biases(user_ids).squeeze()
         item_bias = self.item_biases(item_ids).squeeze()
+
+        # if random.random() < 0.01:
+        #     print('Attention', (user_attention * item_embedding).sum(2).max().data[0])
+        #     print('Softmax', attention.max(1)[0].mean().data[0])
+        #     print('Preference', preference.max(1)[0].mean().data[0])
+        #     print('Prediction', weighted_preference.mean().data[0])
+        #     print('Biases', user_bias.max().data[0], item_bias.max().data[0])
+        # assert False
 
         return weighted_preference + user_bias + item_bias

### 2017-10-28 paper/paper.tex
Start plotting hyperparam search results.

@@
 
-\title{Parameter-efficient multi-modal embeddings for representing diverse user interests}
+\title{Mixture-of-interests models for representing users with diverse interests}
 \author{Maciej Kula}
@@
 
-\keywords{Recommender Systems, Matrix Factorization, Binary Representations}
+\keywords{Recommender Systems, Matrix Factorization, Sequence-based Recommendations}
 
@@
 Most existing recommendation approaches implicitly treat user tastes as unimodal, resulting in an average-of-tastes representations when multiple distinct interests are present.
-We show that appropriately modelling the multi-faceted nature of user tastes through a mixture-of-tastes model leads to large increases in recommendation quality at a very modest cost in model complexity.
+We show that appropriately modelling the multi-faceted nature of user tastes through a mixture-of-tastes model leads to large increases in recommendation quality at a modest cost in model complexity.
 Our result holds both for deep sequence-based and traditional factorization models.

### 2017-10-28 experiment/run_factorization.py

@@ -10,9 +10,10 @@ import torch
 from spotlight.datasets.movielens import get_movielens_dataset
 from spotlight.cross_validation import random_train_test_split
 from spotlight.factorization.implicit import ImplicitFactorizationModel
+# from spotlight.factorization.representations import BilinearNet
 from spotlight.evaluation import mrr_score
 
-from mixture.factorization_representation import MixtureNet
+from mixture.factorization_representation import MixtureNet, NonlinearMixtureNet, BilinearNet, EmbeddingMixtureNet
 
 
 CUDA = torch.cuda.is_available()
@@ -48,20 +49,51 @@ if __name__ == '__main__':
 
     random_state = np.random.RandomState(42)
 
+    hyper = {'batch_size': 192.0 * 10,
+             'embedding_dim': 64,
+             'l2': 2.01514921329504e-11,
+             'learning_rate': 0.004503611222494057 * 1e-2,
+             'loss': 'adaptive_hinge',
+             'n_iter': 10.0,
+             'num_components': 6.0,
+             'type': 'mixture'}
+
+    hyper = {'batch_size': 192,
+             'embedding_dim': 64,
+             'l2': 2.01514921329504e-11,
+             'learning_rate': 0.004503611222494057 * 1e0,
+             'loss': 'adaptive_hinge',
+             'n_iter': 10.0,
+             'num_components': 4.0,
+             'type': 'mixture'}
+    hyper = {'batch_size': 16384.0,
+             'embedding_dim': 32,
+             'l2': 2.0074425248828707e-07,
+             'learning_rate': 0.0048015875347904155,
+             'loss': 'adaptive_hinge',
+             'n_iter': 100.0,
+             'num_components': 3.0,
+             'type': 'mixture'}
+
     train, validation, test = load_data('100K', random_state)
 
-    representation = MixtureNet(train.num_users,
-                                train.num_items,
-                                num_components=4,
-                                embedding_dim=32)
-    model = ImplicitFactorizationModel(loss=loss,
-                                       batch_size=512,
+    representation = EmbeddingMixtureNet(train.num_users,
+                                         train.num_items,
+                                         num_components=int(hyper['num_components']),
+                                         embedding_dim=int(hyper['embedding_dim']))
+    # representation = BilinearNet(train.num_users,
+    #                              train.num_items,
+    #                              embedding_dim=int(hyper['embedding_dim']))
+    model = ImplicitFactorizationModel(loss=hyper['loss'],
[…]

### 2017-10-28 experiment/summarize_results.py

@@ -1,11 +1,14 @@
 from mixture.results import (generate_hyperparameter_table,
                              generate_performance_table,
+                             plot_hyperparam_search,
                              read_results)
 
 
 if __name__ == '__main__':
 
     results = read_results('mixture_results', 'sequence')
+
+    plot_hyperparam_search(results)
 
     print(generate_performance_table(results))
     print(generate_hyperparameter_table(results))

### 2017-10-28 mixture/factorization.py

@@ -10,7 +10,7 @@ from spotlight.evaluation import mrr_score
 from spotlight.factorization.implicit import ImplicitFactorizationModel
 from spotlight.factorization.representations import BilinearNet
 
-from mixture.factorization_representation import MixtureNet
+from mixture.factorization_representation import MixtureNet, NonlinearMixtureNet, EmbeddingMixtureNet
 
 
 CUDA = torch.cuda.is_available()
@@ -19,11 +19,12 @@ CUDA = torch.cuda.is_available()
 def hyperparameter_space():
 
     common_space = {
-        'batch_size': hp.quniform('batch_size', 128, 1024, 64),
-        'learning_rate': hp.loguniform('learning_rate', -6, -3),
+        'batch_size': hp.quniform('batch_size', 512, 16384, 256),
+        'learning_rate': hp.loguniform('learning_rate', -7, -5),
         'l2': hp.loguniform('l2', -25, -9),
-        'embedding_dim': 64,
-        'n_iter': hp.quniform('n_iter', 5, 50, 5),
+        'embedding_dim': 32,
+        'num_components': hp.quniform('num_components', 2, 8, 2),
+        'n_iter': hp.quniform('n_iter', 5, 100, 5),
         'loss': hp.choice('loss', ['adaptive_hinge'])
     }
 
@@ -35,9 +36,16 @@ def hyperparameter_space():
             },
             {
                 'type': 'mixture',
-                'num_components': hp.quniform('num_components', 2, 8, 2),
                 **common_space
             },
+            {
+                'type': 'embedding_mixture',
+                **common_space
+            },
+            # {
+            #     'type': 'nonlinear_mixture',
+            #     **common_space
+            # },
         ])
     }
 
@@ -65,6 +73,16 @@ def get_objective(train, validation, test):
                                         train.num_items,
                                         num_components=int(h['num_components']),
                                         embedding_dim=int(h['embedding_dim']))
+        elif h['type'] == 'nonlinear_mixture':
+            representation = NonlinearMixtureNet(train.num_users,
+                                                 train.num_items,
+                                                 num_components=int(h['num_components']),
+                                                 embedding_dim=int(h['embedding_dim']))
+        elif h['type'] == 'embedding_mixture':
+            representation = EmbeddingMixtureNet(train.num_users,
+                                                 train.num_items,
+                                                 num_components=int(h['num_components']),
+                                                 embedding_dim=int(h['embedding_dim']))
         else:
             raise ValueError('Unknown model type')
 

### 2017-10-28 mixture/factorization_representation.py

@@ -3,7 +3,10 @@ import random
 import torch
 
 import torch.nn as nn
+import torch.nn.init
 import torch.nn.functional as F
+
+from torch.autograd import Variable
 
 from spotlight.layers import ScaledEmbedding, ZeroEmbedding
 
@@ -45,17 +48,19 @@ class MixtureNet(nn.Module):
         self.num_components = num_components
 
         self.user_embeddings = ScaledEmbedding(num_users, embedding_dim)
-
         self.item_embeddings = ScaledEmbedding(num_items, embedding_dim)
 
         self.user_biases = ZeroEmbedding(num_users, 1)
         self.item_biases = ZeroEmbedding(num_items, 1)
 
         self.taste_projection = nn.Linear(embedding_dim,
-                                          embedding_dim * self.num_components)
+                                          embedding_dim * self.num_components, bias=False)
         self.attention_projection = nn.Linear(embedding_dim,
-                                              embedding_dim * self.num_components)
+                                              embedding_dim * self.num_components, bias=False)
 
+        for layer in (self.taste_projection, self.attention_projection):
+            torch.nn.init.xavier_normal(layer.weight, 0.1)
+            
     def forward(self, user_ids, item_ids):
         """
         Compute the forward pass of the representation.
@@ -88,6 +93,138 @@ class MixtureNet(nn.Module):
                           .resize(batch_size,
                                   self.num_components,
                                   embedding_size))
+        user_attention = user_attention #  * user_embedding.unsqueeze(1).expand_as(user_tastes)
+
+        attention = (F.softmax((user_attention *
+                                item_embedding.unsqueeze(1).expand_as(user_attention))
+                               .sum(2)).unsqueeze(2).expand_as(user_attention))
+        weighted_preference = (user_tastes * attention).sum(1)
+
+        dot = (weighted_preference * item_embedding).sum(1)
+
+        user_bias = self.user_biases(user_ids).squeeze()
+        item_bias = self.item_biases(item_ids).squeeze()
+
+        # if random.random() < 0.005:
+            # print('User tastes', user_tastes[0][0].abs().mean().data[0])
+            # print('Tastes', weighted_preference.abs().mean().data[0])
+            # assert False
+        #     print('Attention', (user_attention * item_embedding).sum(2).max().data[0])
+        #     print('Softmax', attention.max(1)[0].mean().data[0])
+        #     print('Preference', preference.max(1)[0].mean().data[0])
[…]

### 2017-10-28 mixture/hyperparameters.py

@@ -19,7 +19,8 @@ def optimize(objective, space, trials_fname=None, max_evals=5):
 
     fmin(objective,
          space=space,
-         algo=hyperopt.rand.suggest,
+         algo=hyperopt.tpe.suggest,
+         #algo=hyperopt.rand.suggest,
          trials=trials,
          max_evals=max_evals)
 

### 2017-10-28 mixture/results.py

@@ -2,9 +2,13 @@ import glob
 import os
 import pickle
 
-from tabulate import tabulate
-
 import numpy as np
+
+import matplotlib.pyplot as plt
+import numpy as np
+import seaborn as sns
+
+from tabulate import tabulate
 
 
 def _get_type(trial):
@@ -18,6 +22,7 @@ def summarize_trials(trials):
                        'mixture', 'mixture2',
                        'linear_mixture', 'diversified_mixture',
                        'diversified_mixture_fixed',
+                       'embedding_mixture',
                        'bilinear'):
         results = [x for x in trials.trials if _get_type(x) == model_type]
         results = sorted(results, key=lambda x: -x['result']['validation_mrr'])
@@ -38,9 +43,21 @@ def _get_best_test_result(results, model_type, filter_fnc=None):
 
     results = [x for x in results.trials
                if _get_type(x) == model_type and filter_fnc(x)]
-    best = sorted(results, key=lambda x: x['result']['validation_mrr'])[-1]
 
-    return best['result']['test_mrr']
+
+    try:
+        best = sorted(results, key=lambda x: x['result']['validation_mrr'])[-1]
+
+        return best['result']['test_mrr']
+    except IndexError:
+        return '-'
+
+
+def _get_test_result_history(results, model_type):
+
+    results = [x for x in results.trials if _get_type(x) == model_type]
+
+    return np.array([x['result']['test_mrr'] for x in results])
 
 
 def read_results(path, variant):
@@ -65,7 +82,8 @@ def generate_performance_table(results):
     rows = []
 
     for (model_name, model) in (('LSTM', 'lstm'),
-                                ('Mixture-LSTM', 'mixture')):
+                                ('Mixture-LSTM', 'mixture'),
+                                ('Linear Mixture-LSTM', 'linear_mixture')):
 
         row = [model_name]
 
@@ -103,3 +121,28 @@ def generate_hyperparameter_table(results):
                     floatfmt='.4f',
                     tablefmt='latex_booktabs')
 
+
+def plot_hyperparam_search(results, max_iter=100):
+
+    fig, (ax0, ax1, ax2) = plt.subplots(1, 3)
+
+    dataset_names = {'10M': 'Movielens 10M',
+                     'amazon': 'Amazon',
+                     'goodbooks': 'Goodbooks-10K'}
+
[…]

### 2017-10-28 mixture/sequence.py

@@ -34,23 +34,23 @@ def hyperparameter_space():
 
     space = {
         'model': hp.choice('model', [
-            # {
-            #     'type': 'lstm',
-            #     **common_space
-            # },
-            # {
-            #     'type': 'mixture',
-            #     **common_space
-            # },
+            {
+                'type': 'lstm',
+                **common_space
+            },
+            {
+                'type': 'mixture',
+                **common_space
+            },
             # {
             #     'type': 'linear_mixture',
             #     **common_space
             # },
-            {
-                'type': 'diversified_mixture_fixed',
-                'diversity_penalty': hp.uniform('diversity_penalty', 0, 2),
-                **common_space
-            },
+            # {
+            #     'type': 'diversified_mixture_fixed',
+            #     'diversity_penalty': hp.uniform('diversity_penalty', 0, 2),
+            #     **common_space
+            # },
         ])
     }
 

### 2017-11-01 experiment/summarize_results.py
Commit results.

@@ -1,6 +1,7 @@
 from mixture.results import (generate_hyperparameter_table,
                              generate_performance_table,
                              plot_hyperparam_search,
+                             plot_hyperparam_bootstrap,
                              read_results)
 
 
@@ -9,6 +10,7 @@ if __name__ == '__main__':
     results = read_results('mixture_results', 'sequence')
 
     plot_hyperparam_search(results)
+    plot_hyperparam_bootstrap(results)
 
     print(generate_performance_table(results))
     print(generate_hyperparameter_table(results))

### 2017-11-01 mixture/results.py

@@ -146,3 +146,53 @@ def plot_hyperparam_search(results, max_iter=100):
 
     fig.tight_layout()
     fig.savefig('hyperparam_search.png')
+
+
+def _bootstrap_max(data, num_samples):
+
+    maxima = []
+
+    for _ in range(num_samples):
+        sample = np.random.choice(data, size=len(data), replace=True)
+        maxima.append(np.max(sample))
+
+    return np.array(maxima)
+
+
+def plot_hyperparam_bootstrap(results, num_samples=10000):
+
+    fig, (ax0, ax1, ax2) = plt.subplots(1, 3)
+
+    dataset_names = {'10M': 'Movielens 10M',
+                     'amazon': 'Amazon',
+                     'goodbooks': 'Goodbooks-10K'}
+
+    for (dataset, ax) in zip(('10M', 'amazon', 'goodbooks'), (ax0, ax1, ax2)):
+
+        baseline = _get_test_result_history(results[dataset], 'lstm')
+        mixture = _get_test_result_history(results[dataset], 'mixture')
+
+        baseline_bootstrap = _bootstrap_max(baseline, num_samples)
+        mixture_bootstrap = _bootstrap_max(mixture, num_samples)
+
+        _, bins = np.histogram(np.concatenate([baseline_bootstrap,
+                                               mixture_bootstrap]),
+                               bins='auto')
+        alpha = 0.75
+        ax.hist(baseline_bootstrap,
+                bins=bins,
+                normed=True,
+                alpha=alpha,
+                label='LSTM')
+        ax.hist(mixture_bootstrap,
+                bins=bins,
+                normed=True,
+                alpha=alpha,
+                label='Mixture-LSTM')
+        ax.set_title(dataset_names[dataset])
+
+        if dataset == 'goodbooks':
+            ax.legend()
+
+    fig.tight_layout()
+    fig.savefig('hyperparam_bootstrap.png')

### 2017-11-01 experiment/summarize_results.py
Add factorization hyperparam plots and results.

@@ -1,16 +1,15 @@
 from mixture.results import (generate_hyperparameter_table,
                              generate_performance_table,
                              plot_hyperparam_search,
-                             plot_hyperparam_bootstrap,
                              read_results)
 
 
 if __name__ == '__main__':
 
-    results = read_results('mixture_results', 'sequence')
+    sequence = read_results('mixture_results', 'sequence')
+    factorization = read_results('mixture_results', 'factorization')
 
-    plot_hyperparam_search(results)
-    plot_hyperparam_bootstrap(results)
+    plot_hyperparam_search(sequence, factorization)
 
-    print(generate_performance_table(results))
-    print(generate_hyperparameter_table(results))
+    print(generate_performance_table(sequence, factorization))
+    # print(generate_hyperparameter_table(results))

### 2017-11-01 mixture/results.py

@@ -11,6 +11,13 @@ import seaborn as sns
 from tabulate import tabulate
 
 
+class Result:
+
+    def __init__(self, trials):
+
+        self.trials = trials
+
+
 def _get_type(trial):
 
     return trial['result']['hyper']['type']
@@ -44,7 +51,6 @@ def _get_best_test_result(results, model_type, filter_fnc=None):
     results = [x for x in results.trials
                if _get_type(x) == model_type and filter_fnc(x)]
 
-
     try:
         best = sorted(results, key=lambda x: x['result']['validation_mrr'])[-1]
 
@@ -76,9 +82,10 @@ def read_results(path, variant):
     return results
 
 
-def generate_performance_table(results):
+def generate_performance_table(sequence, factorization):
 
     headers = ['Model', 'Movielens 10M', 'Amazon', 'Goodbooks-10K']
+    datasets = ('10M', 'amazon', 'goodbooks')
     rows = []
 
     for (model_name, model) in (('LSTM', 'lstm'),
@@ -87,8 +94,20 @@ def generate_performance_table(results):
 
         row = [model_name]
 
-        for dataset in ('10M', 'amazon', 'goodbooks'):
-            mrr = _get_best_test_result(results[dataset], model)
+        for dataset in datasets:
+            mrr = _get_best_test_result(sequence[dataset], model)
+            row.append(mrr)
+
+        rows.append(row)
+
+    for (model_name, model) in (('Bilinear', 'bilinear'),
+                                ('Projection Mixture', 'mixture'),
+                                ('Embedding Mixture', 'embedding_mixture')):
+
+        row = [model_name]
+
+        for dataset in datasets:
+            mrr = _get_best_test_result(factorization[dataset], model)
             row.append(mrr)
 
         rows.append(row)
@@ -122,23 +141,42 @@ def generate_hyperparameter_table(results):
                     tablefmt='latex_booktabs')
 
 
-def plot_hyperparam_search(results, max_iter=100):
+def plot_hyperparam_search(sequence, factorization, max_iter=100):
 
-    fig, (ax0, ax1, ax2) = plt.subplots(1, 3)
+    fig, axes = plt.subplots(2, 3)
+
+    sequence_axes, factorization_axes = (axes[0], axes[1])
 
     dataset_names = {'10M': 'Movielens 10M',
                      'amazon': 'Amazon',
                      'goodbooks': 'Goodbooks-10K'}
 
[…]

### 2017-11-01 paper/paper.tex
Start writing model section.

@@
 \section{Introduction}
+Latent vector-based recommender models commonly represent users with a single latent vector per user \citep{koren2009bellkor, hu2008collaborative}.
+This representation, while simple in formulation and efficient to compute, neglects te fact that users have diverse sub-tastes, and that observed user interactions can be seen as manifestations of several \emph{distinct} tastes or intents.
+These may either be stable facets of the user's preference (liking both horror and documentary movies, or both bluegrass and classical music), context-driven changes (preference for short-form TV content during the week but long-form cinematography during the week-end), or manifestations of phenomena like account sharing, where two (or more) different users, with correspondingly different tastes, share the same user account.
+
+In all these cases, trying to capture the user's taste in a single latent vector is suboptimal.
+Firstly, it may lead to lack of nuance in representing the user, where a dominant taste may overpower more niche ones.
+Secondly it may reduce the quality of item representations, especially when it leads to decreasing the separation in the embedding space between groups of items belonging to multiple tastes or genres.
+For illustration, suppose that documentaries and horror movies are distinct genres, in the sense that most users prefer one or the other: but that there is still a group of users who like both.
+To represent that group of users, the genres' embeddings will have to be separated less cleanly than they would be if the model could express the concept of multiple tastes.
+In general, these problems are similar to that of fitting a unimodal distributional model to bimodal data; for Gaussians, the resulting fitted distribution will be a poor model of the data, and the majority of its density mass will coincide with neither of the true modes.
+
+\section{Model}
+Let $U_i$ be a $m \times k$ matrix representing the $m$ tastes of user $i$, and $A_i$ be a $m \times k$ matrix representing the affinities of each taste for representing particular items.
+The recommendation score for item $j$, represented by a $k times 1$-dimensional embedding vector $e_j$, is then given by
+\begin{equation}
+  r_{ij} = \phi \left(A_ie_j\right) \cdot U_ie_j,
+\end{equation}
+where $\phi$ is the elementwise sigmoid function $\frac{1}{1 + e^{-x}}$.
+
+This gives a mixture of tastes, where the contribution of each individual tastes component is weighted by how competent it is in evaluating item $j$.
+
+How $U_i$ and $A_i$ are obtained differs between the three evaluated models.
+In the Mixture-LSTM model, $U_{it}$ and $A_{it}$ (now indexed by $t$, their position in the sequence of interactions) are linear functions of the hidden state of an LSTM layer trained on the user's previous interactions, $z_{it}$:
+\begin{equation}
+  z_{it} = \mathrm{LSTM}\left(e_{i1}, e_{i2} \ldots e_{it}\right).
+\end{equation}
+Given $z_{it}$, the $m$-th row of $U_{it}$ and $A_{it}$ is given by
+\begin{equation}
+\begin{aligned}
+  u^m_{it} &= W^U_mz_{it} + B^U_m\\
+  a^m_{it} &= W^A_mz_{it} + B^A_m,\\
+\end{aligned}
+\end{equation}
+where $W^U_m$, $W^A_M$, $B^U_m$, and $B^A_m$ are the learned projection matrices. These are common across all users, representing a modest increase in the total number of model parameters.
+
+The Projection Mixture factorization model is similar: an embedding $z_i$ is estimated for each user, and the $U_i$ and $A_i$ matrices are obtained via linear projections:
+\begin{equation}
+\begin{aligned}
+  u^m_{i} &= W^U_mz_{i} + B^U_m\\
+  a^m_{i} &= W^A_mz_{i} + B^A_m.\\
+\end{aligned}
+\end{equation}
+This keeps the number of model parameters small at a potential cost to model capacity.
+In contrast, the Embedding Mixture factorization model embeds $U_i$ and $A_i$ directly.
+This substantially increases the number of model parameters, but may lead to better accuracy.
 
@@
 \midrule
- LSTM         &            0.09 &    0.15  &           0.116 \\
- Mixture-LSTM &            0.1  &    0.189 &           0.136 \\
+ LSTM                &          0.0901 &   0.1502 &          0.1158 \\
+ Mixture-LSTM        &          0.0999 &   0.1889 &          0.1358 \\
+ Linear Mixture-LSTM &          0.0000 &   0.1484 &          0.1163 \\
+ \midrule
+ Bilinear            &          0.0999 &   0.1001 &          0.0738 \\
+ Projection Mixture  &          0.0983 &   0.0698 &          0.0712 \\
+ Embedding Mixture   &          0.0952 &   0.1696 &          0.0853 \\
 \bottomrule
@@
 \end{table*}
+
+\begin{figure*}[h]
+  \caption{Hyperparameter search}
+  \label{fig:hyper}
+  \includegraphics[width=\textwidth]{hyperparam_search.eps}
+\end{figure*}
 

### 2017-11-01 experiment/summarize_results.py

@@ -9,7 +9,7 @@ if __name__ == '__main__':
     sequence = read_results('mixture_results', 'sequence')
     factorization = read_results('mixture_results', 'factorization')
 
-    plot_hyperparam_search(sequence, factorization)
+    plot_hyperparam_search(sequence, factorization, max_iter=25)
 
     print(generate_performance_table(sequence, factorization))
     # print(generate_hyperparameter_table(results))

### 2017-11-01 mixture/results.py

@@ -4,11 +4,17 @@ import pickle
 
 import numpy as np
 
+import matplotlib
+matplotlib.use('Agg')
 import matplotlib.pyplot as plt
 import numpy as np
 import seaborn as sns
 
 from tabulate import tabulate
+
+
+sns.set_style({'font.family': 'serif'})
+sns.set_context('paper')
 
 
 class Result:
@@ -56,7 +62,7 @@ def _get_best_test_result(results, model_type, filter_fnc=None):
 
         return best['result']['test_mrr']
     except IndexError:
-        return '-'
+        return 0.0
 
 
 def _get_test_result_history(results, model_type):
@@ -112,10 +118,12 @@ def generate_performance_table(sequence, factorization):
 
         rows.append(row)
 
-    return tabulate(rows,
-                    headers=headers,
-                    floatfmt='.4f',
-                    tablefmt='latex_booktabs')
+    output = tabulate(rows,
+                      headers=headers,
+                      floatfmt='.4f',
+                      tablefmt='latex_booktabs')
+
+    return output.replace('Bilinear', '\midrule\n Bilinear')
 
 
 def generate_hyperparameter_table(results):
@@ -151,7 +159,8 @@ def plot_hyperparam_search(sequence, factorization, max_iter=100):
                      'amazon': 'Amazon',
                      'goodbooks': 'Goodbooks-10K'}
 
-    for (dataset, ax) in zip(('10M', 'amazon', 'goodbooks'), sequence_axes):
+    for (i, (dataset, ax)) in enumerate(zip(('10M', 'amazon', 'goodbooks'),
+                                            sequence_axes)):
 
         baseline = np.maximum.accumulate(
             _get_test_result_history(sequence[dataset], 'lstm')[:max_iter])
@@ -162,10 +171,15 @@ def plot_hyperparam_search(sequence, factorization, max_iter=100):
         ax.plot(np.arange(len(mixture)), mixture, label='Mixture-LSTM')
         ax.set_title(dataset_names[dataset])
 
-        if dataset == 'goodbooks':
+        if i == 0:
+            ax.set_xlabel('Iterations')
+            ax.set_ylabel('MRR')
+
+        if i == len(factorization_axes):
             ax.legend()
 
-    for (dataset, ax) in zip(('10M', 'amazon', 'goodbooks'), factorization_axes):
+    for (i, (dataset, ax)) in enumerate(zip(('10M', 'amazon', 'goodbooks'),
[…]

### 2017-11-02 paper/paper.tex
Add hyperparameter search section.

@@
 \begin{equation}
-  r_{ij} = \phi \left(A_ie_j\right) \cdot U_ie_j,
+  r_{ij} = \sigma \left(A_ie_j\right) \cdot U_ie_j,
 \end{equation}
-where $\phi$ is the elementwise sigmoid function $\frac{1}{1 + e^{-x}}$.
+where $\sigma$ is the elementwise sigmoid function $\frac{1}{1 + e^{-x}}$.
 
@@
 \begin{equation}
-  z_{it} = \mathrm{LSTM}\left(e_{i1}, e_{i2} \ldots e_{it}\right).
+  z_{it} = \mathrm{LSTM}\left(e_{i1}, e_{i2}, \ldots, e_{it}\right).
 \end{equation}
@@
 This substantially increases the number of model parameters, but may lead to better accuracy.
+
+In all models, the input and output item embeddings $e$ are tied.
 
@@
 Sequences shorter than the maximum sequence length are padded with zeros.
+The models are trained by trying to predict the next item that the user interacts with on the basis of all their prior interactions.
 
@@
 In sequence models, we use the last element of the test interaction sequence as the prediction target; the remaining elements are used to compute the user representation.
-
-We perform extensive hyperparameter optimization across both our proposed models and all baselines to mitigate the danger that our results driven either by models' sensitivity to hyperparameter choices (when a superior model performs poorly because of a poor choice of hyperparameters), or experimenter bias (in favouring the optimization of the novel model).
-We optimize batch size, number of training epochs, the learning rate, L2 regularization weight, the loss function, and (where appropriate) the number of taste mixture components.
 
@@
 \end{itemize}
-For both loss functions, for any known positive user-item interaction pair $(u, i)$, we uniformly sample an implicit negative item $j \in \symUserInteractionSet^-$.
+For both loss functions, for any known positive user-item interaction pair $(i, j)$, we uniformly sample an implicit negative item $k \in \symUserInteractionSet^-$.
 For BPR, the loss for any such triplet is given by
 \begin{equation}
-1 - \sigma\left(r_{ui} - r_{uj}\right),
+1 - \sigma\left(r_{ij} - r_{ik}\right),
 \end{equation}
@@
 \begin{equation}
-\abs{1 - r_{ui} + r_{uj}}_{+}.
+\abs{1 - r_{ij} + r_{ik}}_{+}.
 \end{equation}
-For any $(u, i)$ pair, if the sampled negative item $j$ results in a zero loss (that is, the desired pairwise ordering is not violated), a new negative item is sampled, up to a total of $k$ attempts.
+For any $(i, j)$ pair, if the sampled negative item $k$ results in a zero loss (that is, the desired pairwise ordering is not violated), a new negative item is sampled, up to a maximum number of attempts.
 This leads the model to perform more gradient updates in areas where its ranking performance is poorest.
@@
 Across all of our experiments, the adaptive maximum margin loss consistently outperforms the BPR loss on both baseline and mixture models. We therefore only report results for the adaptive loss.
+
+We perform extensive hyperparameter optimization across both our proposed models and all baselines.
+Our goal is two-fold.
+Firstly, we want to mitigate researcher bias, where more care and attention is devoted to the researcher's proposed model, thus unfairly disadvantaging baseline algorithms in a performance comparison.
+We believe this to be a common phenomenon; its extent is illustrated, in a related domain, by \cite{melis2017state}, who find that standard LSTM architectures, when properly tuned, outperform more recent algorithms in natural laguage modelling tasks.
+Secondly, we wish to understand the extent to which the mixture-of-interests models are \emph{fragile}, in the sense of being highly sensitive to hyperparameter choices.
+Such fragile algorithms are potentially of lesser utility in industry applications, where the engineering cost of tuning in maintaing them may outweigh the accuracy benefits they bring.
+
+We use random search to tune the algorthms used in our experiments.
+We optimize batch size, number of training epochs, the learning rate, L2 regularization weight, the loss function, and (where appropriate) the number of taste mixture components.
+
+\section{Results}
+
+\subsection{Hyperparameter search}
+Figure~\ref{fig:hyper} plots the maximum test MRR obtained by each of our algorithms as a function of the number of elapsed hyperparameter search iterations.
 

### 2017-11-02 mixture/results.py

@@ -175,7 +175,7 @@ def plot_hyperparam_search(sequence, factorization, max_iter=100):
             ax.set_xlabel('Iterations')
             ax.set_ylabel('MRR')
 
-        if i == len(factorization_axes):
+        if i == len(sequence_axes) - 1:
             ax.legend()
 
     for (i, (dataset, ax)) in enumerate(zip(('10M', 'amazon', 'goodbooks'),
@@ -197,7 +197,7 @@ def plot_hyperparam_search(sequence, factorization, max_iter=100):
             ax.set_xlabel('Iterations')
             ax.set_ylabel('MRR')
 
-        if i == len(factorization_axes):
+        if i == len(factorization_axes) - 1:
             ax.legend()
 
     fig.tight_layout()

### 2017-11-03 paper/paper.tex
Start writing related work section.

@@
 
+\section{Related work}
+The idea of moving beyond point embeddings has yielded some interesting results, particularly in the natural language modelling domain.
+
+\citet{vilnis2014word} embed words as Gaussian distributions, rather than point embeddings.
+This improves the resulting representations in two ways.
+Firstly, the resulting representation provides a natural way of expressing uncertainty about the learned representation.
+Secondly, large (and non-spherical) variances aid the representation of polysemous words.
+In the context of recommendations, Gausian embeddings with large variances could be used to represent users with wide-ranging tastes.
+
+\citet{athiwaratkun2017multimodal} extend this idea by representing words a mixtures of Gaussians.
+This allows them to capture polysemy by modelling each word with several distinct, small-variance Gaussian distributions (rather than artificially inflating the variance of a single distribution).
+This allows much clearer separation of distinct meanings.
+This work is closest to the approach presented here.
+
+In the recommender system literature, the idea of using multiple embeddings to represent a user's multiple interests is presented in \citet{weston2013nonlinear}.
+In this approach, the recommendation score of an item for a given user is given by the \emph{maximum} of the dot products of the item embedding and each of the user's embedding vectors.
+This obviates the need for explicit modelling of mixture probabilities, which reduces the number of model parameters and makes evaluation more efficient.
+However, the model is potentially disadvantaged by its inability to model strong \emph{dis}taste for a given class of items.
+
 \section{Model}
@@
 \begin{equation}
-  r_{ij} = \sigma \left(A_ie_j\right) \cdot U_ie_j,
+  r_{ij} = \sigma\left(A_ie_j\right) \cdot U_ie_j,
 \end{equation}
-where $\sigma$ is the elementwise sigmoid function $\frac{1}{1 + e^{-x}}$.
+where $\sigma$ is the elementwise sigmoid function, $\sigma\left(A_ie_j\right)$ gives the mixture probabilities, and $U_ie_j$ the recommendation scores output by each mixture component.
+We assume identity variance matrices for all mixture components.
 

### 2017-11-03 paper/paper.tex
Add number of mixtures sensitivity table.

@@
-\documentclass[sigchi]{acmart}
+\documentclass[sigconf]{acmart}
 
 \usepackage{booktabs}
+\usepackage{subcaption}
 \usepackage{commath}
@@
 
+In this paper, we propose and evaluate representing users as mixtures of several distinct tastes, represented by distinct taste vectors.
+Each of the taste vectors is coupled with an attention vector, describing how competent it is at evaluating any given item.
+The user's preference is then modelled as a weighted average of all the user's tastes, with the weights given by how relevant each taste is to evaluating a given item.
+
+We apply this model to both traditional implicit feedback factorization models, and to more recent models using recurrent neural networks.
+In both cases, our mixture-of-interests models handily outperform single-representation models on standard ranking quality metrics.
+In the case of deep recurrent sequence models, this improvement is achieved at a very modest cost to to the model complexity, making our model a straightforward improvement over existing methods.
+
 \section{Related work}
@@
 \end{equation}
-where $W^U_m$, $W^A_M$, $B^U_m$, and $B^A_m$ are the learned projection matrices. These are common across all users, representing a modest increase in the total number of model parameters.
+where $W^U_m$, $W^A_M$, $B^U_m$, and $B^A_m$ are the learned projection matrices.
+These are common across all users, representing a modest increase in the total number of model parameters.
+Note that the LSTM network only needs to be run once to obtain the full user representation.
 
@@
 
-\begin{table*}[htbp]
+\begin{table}[htbp]
 \begin{threeparttable}
@@
 \end{threeparttable}
-\end{table*}
+\end{table}
+
+\begin{table}
+\caption{Effect of number of mixture components}
+\label{tab:nummixtures}
+\begin{subtable}{\columnwidth}
+\caption{Sequence models}
+\begin{tabular}{rrrr}
+\toprule
+   Mixture components &   Movielens 10M &   Amazon &   Goodbooks-10K \\
+\midrule
+                    2 &          0.0921 &   0.1538 &          0.1262 \\
+                    4 &          0.0999 &   0.1689 &          0.1304 \\
+                    6 &          0.0991 &   0.1889 &          0.1358 \\
+                    8 &          0.0982 &   0.1865 &          0.1327 \\
+\bottomrule
+\end{tabular}
+\end{subtable}
+\begin{subtable}{\columnwidth}
+\caption{Factorization models}
+\begin{tabular}{rrrr}
+\toprule
+   Mixture components &   Movielens 10M &   Amazon &   Goodbooks-10K \\
+\midrule
+                    2 &          0.0952 &   0.1343 &          0.0797 \\
+                    4 &          0.0952 &   0.1583 &          0.0840 \\
+                    6 &          0.0913 &   0.1612 &          0.0811 \\
+                    8 &          0.0907 &   0.1696 &          0.0853 \\
+\bottomrule
+\end{tabular}
+\end{subtable}\end{table}
 

### 2017-11-03 experiment/summarize_results.py

@@ -12,4 +12,4 @@ if __name__ == '__main__':
     plot_hyperparam_search(sequence, factorization, max_iter=25)
 
     print(generate_performance_table(sequence, factorization))
-    # print(generate_hyperparameter_table(results))
+    print(generate_hyperparameter_table(sequence, factorization))

### 2017-11-03 mixture/results.py

@@ -126,27 +126,49 @@ def generate_performance_table(sequence, factorization):
     return output.replace('Bilinear', '\midrule\n Bilinear')
 
 
-def generate_hyperparameter_table(results):
+def generate_hyperparameter_table(sequence, factorization):
 
     headers = ['Mixture components', 'Movielens 10M', 'Amazon', 'Goodbooks-10K']
-    rows = []
 
-    for num_components in (2, 4, 6, 8):
+    outputs = []
 
-        row = [num_components]
+    for (model, results) in (('mixture', sequence), ('embedding_mixture', factorization)):
 
-        filter_fnc = lambda x: x['result']['hyper'].get('num_components') == num_components
-        
-        for dataset in ('10M', 'amazon', 'goodbooks'):
-            mrr = _get_best_test_result(results[dataset], 'mixture', filter_fnc)
-            row.append(mrr)
+        rows = []
 
-        rows.append(row)
+        for num_components in (2, 4, 6, 8):
 
-    return tabulate(rows,
-                    headers=headers,
-                    floatfmt='.4f',
-                    tablefmt='latex_booktabs')
+            row = [num_components]
+
+            def filter_fnc(x):
+                return x['result']['hyper'].get('num_components') == num_components
+
+            for dataset in ('10M', 'amazon', 'goodbooks'):
+                mrr = _get_best_test_result(results[dataset], model, filter_fnc)
+                row.append(mrr)
+
+            rows.append(row)
+
+        output = tabulate(rows,
+                          headers=headers,
+                          floatfmt='.4f',
+                          tablefmt='latex_booktabs')
+        outputs.append(output)
+
+    sequence_table, factorization_table = outputs
+
+    return ('\\begin{table}\n'
+            '\\caption{Effect of number of mixture components}\n'
+            '\\label{tab:nummixtures}\n'
+            '\\begin{subtable}{\\columnwidth}\n'
+            '\\caption{Sequence models}\n'
+            '%s\n'
+            '\\end{subtable}\n'
+            '\\begin{subtable}{\\columnwidth}\n'
+            '\\caption{Factorization models}\n'
+            '%s\n'
+            '\\end{subtable}'
+            '\\end{table}' % (sequence_table, factorization_table))
 
 
 def plot_hyperparam_search(sequence, factorization, max_iter=100):
[…]

### 2017-11-03 paper/paper.tex
Plot formatting improvements.

@@
 \subsection{Datasets}
-For our experiments, we use the following datasets.
+We use the following datasets in out experiments.
 
-\paragraph{Movielens 10M} A dataset of 10 million ratings across 10,000 movies and 72,000 users \citep{harper2016movielens}.
-\paragraph{Goodbooks} A dataset of 6 million ratings across 53,000 users and 10,000 most popular books from the Goodbooks online book recommendation and sharing service \citep{goodbooks2017}.
-\paragraph{Amazon} A dataset of ratings and reviews gathered from the Amazon online shopping service.
+\begin{enumerate}
+\item Movielens 10M: dataset of 10 million movie ratings across 10,000 movies and 72,000 users \citep{harper2016movielens}.
+\item Goodbooks: dataset of 6 million ratings across 53,000 users and 10,000 most popular books from the Goodbooks online book recommendation and sharing service \citep{goodbooks2017}.
+\item Amazon: dataset of ratings and reviews gathered from the Amazon online shopping service \citep{leskovec2007dynamics}.
 After pruning users and items with fewer than 10 ratings, the dataset contains approximately 4 million ratings from 100,000 users over 114,000 items.
+\end{enumerate}
 
@@
 \toprule
-   Mixture components &   Movielens 10M &   Amazon &   Goodbooks-10K \\
+   Components &   Movielens 10M &   Amazon &   Goodbooks-10K \\
 \midrule
@@
   \label{fig:hyper}
-  \includegraphics[width=\textwidth]{hyperparam_search.eps}
+  \includegraphics[width=0.8\textwidth]{hyperparam_search.eps}
 \end{figure*}
-
-%% \section{Introduction}
-%% Industry ranking and recommendation systems need to scale to tens or hundreds of millions of items and users.
-As the number of items available to be recommended grows large, efficiently ranking the item catalogue to produce top-ranked recommendations becomes increasingly challenging.
-With large catalogues, ranking latency (in on-line settings) and compute and storage costs (in pre-compute settings) place significant constraints on the design of the entire recommender system.
-
-%% In an on-line setting, where recommendation latency directly impacts the user experience, system designers have to work within a strict latency budget.
-Respecting that budget often means trading off model accuracy for speed, either via smaller and less expressive models (with lower-dimensionality representations) or through aggressive use of heuristics to exclude large classes of candidate items from scoring.
-For example, Pinterest uses a combination of heuristics and candidate-generation models to select approximately 1000 pins, from billions of available pins, as input to its related pins ranking system \citep{liu2017related}.
-Similarly, approximate nearest neighbour search techniques are used at YouTube \citep{covington2016deep} to reduce the number of candidate videos that need to be scored for each user.
-
-%% The challenge of low-latency scoring leads many systems to rely on a pre-compute system, where recommendations are calculated in advance and stored for serving at a later time.
-The primary disadvantage of such a system is its inability to react dynamically to user actions (such as new interactions that allow the recommendations to be refined) and recommendation context (location, time of day or year).
-This leads to often complex architectures that mix fairly simple on-line models, computationally expensive offline models, and intermediate complexity near-line algorithms where model updating is important \citep{amatriain2013big}.
-
-%% Additionally, while a pre-compute solution removes model prediction from the critical path, it still requires substantial investment in computational and storage resources to calculate and cache the predictions of the model.
-Often the amount of computation required makes timely updating of the cached results impossible.
-At Pinterest, pre-calculation of related pins results was so demanding that ranking of different segments of the catalogue had to be staggered in time, leading to stale results and reduced ease of development \citep[Section 5.5]{liu2017related}.
-\citet{koenigstein2012efficient} estimate that computing recommendations for the Yahoo!
-Music dataset \citep{dror2011yahoo} with a 50-dimensional latent factor model woud take over 135 hours.
-
-%% To alleviate these constraints in both online and offline settings, we evaluate the use of binary item and user representations in learning-to-rank matrix factorization models.
-Following the approach of \citet{rastegari2016xnor}, we estimate binary user and item representations that are several times faster to score and requiring a fraction of the memory (for the same latent dimensionality).
-Unfortunately, they are substantially less accurate than standard learning-to-rank approaches.
-The resulting speed-accuracy trade-offs favour simply reducing the latent dimensionality of traditional models when evaluation speed is desired.
-
-%% We speculate that binary representations are more useful in case of more complex models (such as deep convolutional models) where a single prediction requires many more floating operations than relatively compact bilinear recommender models.
-
-%% \section{Binary latent representations}
-%% \label{sec:approach}
-%% \citet{rastegari2016xnor} introduce XNOR-Networks, an approach that combines 1-bit quantization of fully-connected and convolutional layers with representing dot products through scaled XNOR and bitcounting operations.
-
-%% We adapt their approach to the recommendation task by learning binary versions of user and item vectors within a learning-to-rank factorization model.
-This allows us to use binary instead of floating point operations for computing recommendation rankings, which has significant speed and memory use advantages.
-In a real-valued $n$-dimensional factorization model, item and user representations take $4n$ bytes of memory, and computing a single prediction takes $2n$ floating point operations.
-In an equivalent binary model, $\frac{3}{32}n$ binary operations are required for each prediction (XOR, negation, and bitcounting), and the representations take $\frac{1}{8}n$ bytes of memory.
-Naive calculations suggest that, for the same latent dimensionality, a binary model would be over 20 times faster to score, and take less then 5\% of memory to store its representations.
-
-%% Naturally, using binary instead of real-valued representations entails a loss of representational power, and the resulting models can be expected to be less accurate.
-Nevertheless, binary models may still be very useful in a production system, for two reasons.
-Firstly, they may offer the system designer a more favourable trade-off between speed and accuracy than simply changing the number of latent dimensions in a real-valued model.
-If latency is the binding constraint, switching to a binary model may offer better or comparable speed with better ranking performance.
-If accuracy is the priority, a binary model may offer equivalent ranking performance while being faster to score.
-Secondly, many production systems resort to heavy use of heuristics when selecting candidates to be scored by their ranking systems (as scoring all candidates would be infeasible).
-To the extent that these heuristics are suboptimal, replacing them with a binary representation model would improve the overall accuracy of the system.
-
-%% To quantify the trade-offs involved, we construct and fit both a real-valued and a binary version of simple implicit learning-to-rank factorization model.
-We introduce the notation and the models below.
-
-%% Let $\symUserSet$ be the set of users, and $\symItemSet$ be the set of items.
-Each user interacts with a number of items $\symUserInteractionSet^+$; the set of all remaining items is denoted by $\symUserInteractionSet^-$.
-
-%% In a traditional real-valued latent factor model, the prediction $r_{ui}$ for any user-item interaction pair $(u, i) \in \symUserSet \times \symItemSet$ is given by
-%% \begin{equation}
-%% r_{ui} = \vec{u}_u \cdot \vec{i}_i + b_u + b_i
-%% \end{equation}
-%% where $\vec{u}_u$ denotes the user and $\vec{i}_i$ the item $n$-dimensional latent vectors, and $b_u$ and $b_i$ the user and item biases.
-
-%% \citet{rastegari2016xnor} show that the dot product of two real-valued vectors can be approximated in the binary domain by (1) binarizing the input vectors using the sign function (so that the results lie in $\{1, -1\}$), (2) taking their dot product, and (3) scaling the result by the average magnitude of the vectors' elements.
-Letting
-%% \begin{equation}
-%% \alpha = \frac{1}{n}\norm{\vec{i}_{i}}_{\ell1}
-%% \end{equation}
-%% and
-%% \begin{equation}
-%% \beta = \frac{1}{n}\norm{\vec{u}_{u}}_{\ell1}
-%% \end{equation}
-%% represent the scaling factors, the approximation is given by
-%% \newcommand\binaryApproximation{ \alpha \beta\left(\sign(\vec{u}_u) \cdot \sign(\vec{i}_i)\right)}
-%% \begin{equation}
-%% \vec{u}_u \cdot \vec{i}_i \approx \binaryApproximation
-%% \end{equation}
-%% Applying this to the latent-factor model, the prediction for a user-item pair is given by
-%% \begin{equation}
-%% r_{ui} = \binaryApproximation + b_u + b_i
-%% \end{equation}
-%% Note that the scaling factors $\alpha$ and $\beta$ need to be stored (in addition to the latent representations and biases) to compute predictions in the binary model.
-Their use also implies a constant cost of two floating-point multiplications per dot product when using binary representations.
-
-%% The model is trained using backpropagation, and so we use a smooth approximation to the derivative of the sign function to facilitate training. Following \citet{courbariaux2016binarized}, we define
-%% \begin{equation}
-%%   \frac{d\sign}{dw}=
-%%   \begin{cases}
-%%     1 & \text{if}\ \abs{w} \leq 1 \\
-%%     0 & \text{otherwise}.
-%%   \end{cases}
-%% \end{equation}
-
-%% We use two loss functions to train the model:
-%% \begin{itemize}
-%% \item Bayesian personalised ranking (BPR, \citet{rendle2009bpr}), and
-%% \item adaptive sampling maximum margin loss, following \citet{weston2011wsabie}.
-%% \end{itemize}
-%% For both loss functions, for any known positive user-item interaction pair $(u, i)$, we uniformly sample an implicit negative item $j \in \symUserInteractionSet^-$.
-For BPR, the loss for any such triplet is given by
-%% \begin{equation}
-%% 1 - \sigma\left(r_{ui} - r_{uj}\right),
-%% \end{equation}
-%% where $\sigma$ denotes the sigmoid function.
-%% The adaptive sampling loss is given by
-%% \begin{equation}
-%% \abs{1 - r_{ui} + r_{uj}}_{+}.
-%% \end{equation}
-%% For any $(u, i)$ pair, if the sampled negative item $j$ results in a zero loss (that is, the desired pairwise ordering is not violated), a new negative item is sampled, up to a total of $k$ attempts.
-This leads the model to perform more gradient updates in areas where its ranking performance is poorest.
-
-%% For the purposes of this paper, we focus on a simple bilinear collaborative filtering model, and do not use any external metadata information or model the sequential nature of the data.
-However, the binary dot product approach generalizes beyond simple factorization models, and can be applied as a component of any model whose final scoring step involves a dot product between user and item representations.
-In particular, models using recurrent \citep{hidasi2015session} or convolutional \citep{lynch2015images} item or user representations can easily be augmented to use binary dot products in the final ranking stages.
-
-%% \section{Experiments}
-%% \begin{table*}[htbp]
-%% \begin{threeparttable}
-%% \caption{Movielens 1M results}
-%% \label{tb:results}
-%% \centering
-%% \begin{tabular}{rrrrrrrr}
-%% \toprule
-%% Dimension &   MRR & Binary MRR & MRR ratio\tnote{1} &   PPMS\tnote{2} & Binary PPMS & PPMS ratio\tnote{3} & Memory use ratio\tnote{4} \\
-%% \midrule
-%%        32 & 0.077 &      0.059 &     0.768 & 87,758 &     264,146 &      3.010 &            0.091 \\
-%%        64 & 0.075 &      0.066 &     0.878 & 45,992 &     220,833 &      4.802 &            0.062 \\
-%%       128 & 0.076 &      0.071 &     0.935 & 23,870 &     166,252 &      6.965 &            0.047 \\
-%%       256 & 0.078 &      0.071 &     0.908 & 12,284 &     190,131 &     15.477 &            0.039 \\
-%%       512 & 0.080 &      0.073 &     0.912 &   6074 &     122,637 &     20.190 &            0.035 \\
-%%      1024 & 0.080 &      0.075 &     0.936 &   3056 &      61,301 &     20.056 &            0.033 \\
-%% \bottomrule
-%% \end{tabular}
-%% \begin{tablenotes}
-%% \small{
-%% \item[1] Ratio of binary model MRR to real-valued model MRR
-%% \item[2] Predictions Per Millisecond: how many items can be scored per millisecond
-%% \item[3] Ratio of binary PPMS to real-valued PPMS
-%% \item[4] Ratio of memory required to store binary vs. real-valued parameters
-%% }
-%% \end{tablenotes}
-%% \end{threeparttable}
-%% \end{table*}
-%% To assess the accuracy-speed trade-offs enabled by the XNOR approach, we conduct an experiment on the Movielens 1M dataset \citep{harper2016movielens}.
-The dataset contains 1 million ratings from 6000 users on 4000 movies.
-Because the computational speed improvements offered by the binary representations are linear in the size of the catalogue, the results on this dataset should be indicative of results that can be obtained on larger datasets.
-At the same time, the dataset's small size enables us to easily test hyperparameter configurations and estimate models with high latent dimensions.
-
-%% \subsection{Experimental setup}
-%% We randomly divide the dataset into a training, test, and validation set.
-In order to build a picture of the accuracy-speed trade-offs afforded by real-valued and binary latent models, we explicitly build models for between 32 and 1024 latent dimensions.
-For every latent dimensionality, we conduct a random search over the hyperparameter space, and pick optimal initial learning rate, loss function, L2 penalty, minibatch size and number of training epochs for each algorithm based on their performance on the test set.
-The final results are obtained from ranking interactions from the validation set.
-We use mean reciprocal rank (MRR) as a measure of ranking quality.
-
-%% Benchmark results are measured on an Intel Xeon E5-2686v4 CPU by running a 500 repetitions of scoring 100,000 items and averaging the results.
-
-%% \subsection{Implementation}
-%% The model is implemented in PyTorch and trained using the nVidia K40 GPUs.
-The implementation is only marginally more complex than a standard learning-to-rank model, and is accomplished by simply replacing the user-item vector dot product operation with its binary counterpart.
-The change results in a small increase in training time.
-
-%% During training, the embedding parameters are stored and updated as single precision floating point values.
-Similarly, the XNOR dot product is carried out using floats in $\{1, -1\}$.
-The initial learning rate, loss function, L2 penalty, minibatch size and number of training epochs are treated as model hyperparameters.
-All models are trained using the Adam training rate schedule \citep{kingma2014adam}.
-
-%% The prediction code runs on the CPU and is implemented in C using Intel X86 AVX2 SIMD intrinsics.
-SIMD (Single Instruction Multiple Data) instructions allow the CPU to operate on multiple pieces of data in parallel, achieving significant speedups over the scalar version.
-We use explicit intrinsics rather than compiler autovectorization to ensure that neither the real-valued nor the binary prediction code is unfairly disadvantaged by the quality of compiler autovectorization.
-
-%% The real-valued prediction code is implemented using 8-float wide fused multiply-add instructions (\texttt{\_mm256\_fmadd\_ps}).
-In the binary version, the real-valued embedding parameters used in training are discarded, and the derived 1-bit weights are packed into 32-bit integer buffers.
-The XNOR dot product is implemented using 8-integer wide XOR operations (256 binary weights are processed at a time), followed by a popcount instruction to count the number of on bits in the result.
-We use the \texttt{libpopcnt} \citep{mula2016faster} library for the bit counting operations.
-
-%% Both versions use 32-bit aligned input data to utilise aligned SIMD register load instructions. The code is compiled using GCC 4.8.4 for the AVX2-enabled Broadwell architecture.
-
-%% The source code for both model training and prediction is available on Github at \url{https://github.com/maciejkula/binge}.
-
-%% \subsection{Results}
-%% \label{sec:results}
-%% Table \ref{tb:results} summarises the results of the Movielens 1M experiment.
-Each row presents the best MRR (mean reciprocal rank) results for both real-valued and binary models of a given dimensionality.
-MRR ratio denotes the fraction of the real-valued MRR the binary model achieves for that dimensionality; PPMS (predictions per millisecond) ratio denotes how many more predictions the binary model can compute per millisecond.
-
-%% Results on scoring speed and memory use broadly confirm the naive calculations from section \ref{sec:approach}, converging to over 20 times faster scoring at around 3\% of memory as latent dimensionality increases.
-Memory savings plateau as latent dimensionality increases and storing the binary scaling factors becomes less important.
-
-%% As expected, for the same dimensionality a binary model achieves lower ranking accuracy.
-On average, the accuracy loss when moving from a continuous to a binary model of the same latent dimensionality is around 11\%, varying between 6\% and 23\%.
-
-%% Unfortunately, while continuous models retain good accuracy as latent dimensionality decreases, binary models' representational power sharply deteriorates.
-Moving from the 1024 to 32 dimensions in the continuous model implies a 29 times increase in prediction speed at the expense of a modest 4\% decrease in accuracy.
-This compares favourably to switching to binary representations: moving to a 1024-binary dimensional representation implies a sharper accuracy drop at 6\% in exchange for a smaller 20 times increase in prediction speed.
-Moving to 32 dimensions yields a further speed gains at 86 times, but at the cost of a considerable loss of accuracy at 26\%.
-
-%% The results suggest that latent recommender models have good representational power even at low latent dimensionalities.
-This advantage is lost when using binary embeddings, which need to be high-dimensional to achieve comparable accuracy.
-At the same time, binary representations' speed advantage is only evident at high dimensions: at low dimensions, only floating-point operations can use enjoy the advantages of SIMD operations, and the fixed overhead of binary scaling factors $\alpha$ and $\beta$ constitutes a larger proportion of the total computational cost.
-We conjecture that the attractiveness of binary representations is tightly coupled with high-dimensional models where a single prediction requires many floating point operations, such as convolutional neural networks.
-Relatively compact latent factor models do not fall into this category.
-
-%% \section{Alternative approaches}
-%% Given the limited success of binary embedding, a more promising approach is offered by existing work that focuses on reducing the number of dot products which need to be performed in order to retrieve top recommendations.
-\citet{koenigstein2012efficient} introduces an exact branch-and-bound based on metric trees as well as an approximate algorithm that clusters users and serves recommendations computed for the cluster centers.
-\citet{shrivastava2014asymmetric} extend well-known \citep{indyk1998approximate} locality-sensitive hashing techniques to maximum inner product search (MIPS) through asymmetric transformations of the query and candidate vectors.
-
-%% The method that is most closely related to the model we present in this paper is \citet{Shen_2015_ICCV}, who expand on MIPS by learning asymmetric binary hash functions.
-The hash functions are trained to minimize the L2 norm between the inner product matrix of the original input vectors and the inner product matrix of their binary representations.
-
-%% \section{Conclusion}
-%% While prediction latency is a pressing concern for many recommender systems, we find that the already compact latent factor models do not stand to benefit from using binary latent representations.
-The sharp drop in representational power exhibited by binary embeddings makes other approaches (such as exact \citep{koenigstein2012efficient} and approximate \citep{shrivastava2014asymmetric} maximum inner product search) more promising avenues when optimizing large-scale ranking.
 

### 2017-11-03 mixture/results.py

@@ -13,8 +13,8 @@ import seaborn as sns
 from tabulate import tabulate
 
 
-sns.set_style({'font.family': 'serif'})
-sns.set_context('paper')
+sns.set_style('ticks', {'font.family': 'serif', 'axes.linewidth': 0.8})
+sns.set_context('paper', font_scale=0.8)
 
 
 class Result:
@@ -128,7 +128,7 @@ def generate_performance_table(sequence, factorization):
 
 def generate_hyperparameter_table(sequence, factorization):
 
-    headers = ['Mixture components', 'Movielens 10M', 'Amazon', 'Goodbooks-10K']
+    headers = ['Components', 'Movielens 10M', 'Amazon', 'Goodbooks-10K']
 
     outputs = []
 

### 2017-11-03 experiment/summarize_results.py
Add dataset stats table.

@@ -1,4 +1,5 @@
-from mixture.results import (generate_hyperparameter_table,
+from mixture.results import (generate_dataset_table,
+                             generate_hyperparameter_table,
                              generate_performance_table,
                              plot_hyperparam_search,
                              read_results)
@@ -13,3 +14,4 @@ if __name__ == '__main__':
 
     print(generate_performance_table(sequence, factorization))
     print(generate_hyperparameter_table(sequence, factorization))
+    print(generate_dataset_table())

### 2017-11-03 mixture/results.py

@@ -12,6 +12,9 @@ import seaborn as sns
 
 from tabulate import tabulate
 
+from spotlight.datasets.movielens import get_movielens_dataset
+from spotlight.datasets.goodbooks import get_goodbooks_dataset
+from spotlight.datasets.amazon import get_amazon_dataset
 
 sns.set_style('ticks', {'font.family': 'serif', 'axes.linewidth': 0.8})
 sns.set_context('paper', font_scale=0.8)
@@ -224,3 +227,29 @@ def plot_hyperparam_search(sequence, factorization, max_iter=100):
 
     fig.tight_layout()
     fig.savefig('hyperparam_search.eps')
+
+
+def generate_dataset_table():
+
+    headers = ['Dataset', 'Users', 'Items', 'Interactions', 'Density']
+
+    rows = []
+
+    for name, dataset in (('Movielens 10M', get_movielens_dataset('10M')),
+                          ('Amazon', get_amazon_dataset()),
+                          ('Goodbooks-10K', get_goodbooks_dataset())):
+
+        row = [
+            name,
+            '{:0,}'.format(dataset.num_users),
+            '{:0,}'.format(dataset.num_items),
+            '{:0,}'.format(len(dataset)),
+            len(dataset) / dataset.num_users / dataset.num_items
+        ]
+
+        rows.append(row)
+
+    return tabulate(rows,
+                    headers=headers,
+                    floatfmt='.4f',
+                    tablefmt='latex_booktabs')

### 2017-11-04 paper/paper.tex
Add dataset stats table.
Improve table automation.

@@
 \usepackage{threeparttable}
+\usepackage{caption}
 
@@
 
+\begin{table}
+  \caption{Dataset statistics}
+  \label{tab:datasets}
+  \input{includes/datasets.tex}
+\end{table}
+
 \subsection{Experimental setup}
@@
 \subsection{Hyperparameter search}
-Figure~\ref{fig:hyper} plots the maximum test MRR obtained by each of our algorithms as a function of the number of elapsed hyperparameter search iterations.
+Figure~\ref{fig:hyper} plots the maximum test MRR achieved by each algorithm as a function of the number of elapsed hyperparameter search iterations.
+Both baseline and mixture models benefit from hyperparameter tuning.
+All algorithms converge to their optimum performance relatively quickly, suggesting a degree of robustness to hyperparameter choices.
+Mixture-LSTM and Embedding Mixture models quickly outperform their baseline counterparts, and maintain a stable performance lead thereafter.
+This lends support to our belief that the mixture models' superior accuracy reflects their greater capacity to model the recommendation problem well, rather than being an artifact of the experimental procedure or researcher bias.
+
+\subsection{Number of taste components}
+We summarize the effect of increasing the number of taste mixture components in Table~\ref{tab:nummixtures}.
+By and large, there is a dose-response relationship between the number of mixtures and recommendation quality: being able to represents more distinct user tasts yields better results.
 
@@
 \label{tb:results}
-\begin{tabular}{lrrr}
-\toprule
- Model               &   Movielens 10M &   Amazon &   Goodbooks-10K \\
-\midrule
- LSTM                &          0.0901 &   0.1502 &          0.1158 \\
- Mixture-LSTM        &          0.0999 &   0.1889 &          0.1358 \\
- Linear Mixture-LSTM &          0.0000 &   0.1484 &          0.1163 \\
- \midrule
- Bilinear            &          0.0999 &   0.1001 &          0.0738 \\
- Projection Mixture  &          0.0983 &   0.0698 &          0.0712 \\
- Embedding Mixture   &          0.0952 &   0.1696 &          0.0853 \\
-\bottomrule
-\end{tabular}
+\input{includes/performance.tex}
 \begin{tablenotes}
@@
 
-\begin{table}
-\caption{Effect of number of mixture components}
-\label{tab:nummixtures}
-\begin{subtable}{\columnwidth}
-\caption{Sequence models}
-\begin{tabular}{rrrr}
-\toprule
-   Components &   Movielens 10M &   Amazon &   Goodbooks-10K \\
-\midrule
-                    2 &          0.0921 &   0.1538 &          0.1262 \\
-                    4 &          0.0999 &   0.1689 &          0.1304 \\
-                    6 &          0.0991 &   0.1889 &          0.1358 \\
-                    8 &          0.0982 &   0.1865 &          0.1327 \\
-\bottomrule
-\end{tabular}
-\end{subtable}
-\begin{subtable}{\columnwidth}
-\caption{Factorization models}
-\begin{tabular}{rrrr}
-\toprule
-   Components &   Movielens 10M &   Amazon &   Goodbooks-10K \\
-\midrule
-                    2 &          0.0952 &   0.1343 &          0.0797 \\
-                    4 &          0.0952 &   0.1583 &          0.0840 \\
-                    6 &          0.0913 &   0.1612 &          0.0811 \\
-                    8 &          0.0907 &   0.1696 &          0.0853 \\
-\bottomrule
-\end{tabular}
-\end{subtable}\end{table}
+\input{includes/hyperparameters.tex}
 
 \begin{figure*}[h]
-  \caption{Hyperparameter search}
+  \captionsetup{width=.8\linewidth}
+  \caption{Maximum test MRR vs number of hyperparameter search iterations. Sequence-based models in the top row; factorization-based models in the bottom row.}
   \label{fig:hyper}
-  \includegraphics[width=0.8\textwidth]{hyperparam_search.eps}
+  \includegraphics[width=0.8\textwidth]{includes/hyperparam_search.pdf}
 \end{figure*}

### 2017-11-04 experiment/summarize_results.py

@@ -1,3 +1,5 @@
+import os
+
 from mixture.results import (generate_dataset_table,
                              generate_hyperparameter_table,
                              generate_performance_table,
@@ -7,11 +9,19 @@ from mixture.results import (generate_dataset_table,
 
 if __name__ == '__main__':
 
+    dest_dir = os.path.abspath('paper/includes')
+
     sequence = read_results('mixture_results', 'sequence')
     factorization = read_results('mixture_results', 'factorization')
 
-    plot_hyperparam_search(sequence, factorization, max_iter=25)
+    fig = plot_hyperparam_search(sequence, factorization, max_iter=25)
+    fig.savefig(os.path.join(dest_dir, 'hyperparam_search.pdf'))
 
-    print(generate_performance_table(sequence, factorization))
-    print(generate_hyperparameter_table(sequence, factorization))
-    print(generate_dataset_table())
+    with open(os.path.join(dest_dir, 'performance.tex'), 'w') as fle:
+        fle.write(generate_performance_table(sequence, factorization))
+
+    with open(os.path.join(dest_dir, 'hyperparameters.tex'), 'w') as fle:
+        fle.write(generate_hyperparameter_table(sequence, factorization))
+
+    with open(os.path.join(dest_dir, 'datasets.tex'), 'w') as fle:
+        fle.write(generate_dataset_table())

### 2017-11-04 mixture/results.py

@@ -93,13 +93,12 @@ def read_results(path, variant):
 
 def generate_performance_table(sequence, factorization):
 
-    headers = ['Model', 'Movielens 10M', 'Amazon', 'Goodbooks-10K']
+    headers = ['Model', 'Movielens 10M', 'Amazon', 'Goodbooks']
     datasets = ('10M', 'amazon', 'goodbooks')
     rows = []
 
     for (model_name, model) in (('LSTM', 'lstm'),
-                                ('Mixture-LSTM', 'mixture'),
-                                ('Linear Mixture-LSTM', 'linear_mixture')):
+                                ('Mixture-LSTM', 'mixture')):
 
         row = [model_name]
 
@@ -131,7 +130,7 @@ def generate_performance_table(sequence, factorization):
 
 def generate_hyperparameter_table(sequence, factorization):
 
-    headers = ['Components', 'Movielens 10M', 'Amazon', 'Goodbooks-10K']
+    headers = ['Components', 'Movielens 10M', 'Amazon', 'Goodbooks']
 
     outputs = []
 
@@ -182,7 +181,7 @@ def plot_hyperparam_search(sequence, factorization, max_iter=100):
 
     dataset_names = {'10M': 'Movielens 10M',
                      'amazon': 'Amazon',
-                     'goodbooks': 'Goodbooks-10K'}
+                     'goodbooks': 'Goodbooks'}
 
     for (i, (dataset, ax)) in enumerate(zip(('10M', 'amazon', 'goodbooks'),
                                             sequence_axes)):
@@ -226,7 +225,8 @@ def plot_hyperparam_search(sequence, factorization, max_iter=100):
             ax.legend()
 
     fig.tight_layout()
-    fig.savefig('hyperparam_search.eps')
+
+    return fig
 
 
 def generate_dataset_table():
@@ -237,7 +237,7 @@ def generate_dataset_table():
 
     for name, dataset in (('Movielens 10M', get_movielens_dataset('10M')),
                           ('Amazon', get_amazon_dataset()),
-                          ('Goodbooks-10K', get_goodbooks_dataset())):
+                          ('Goodbooks', get_goodbooks_dataset())):
 
         row = [
             name,

### 2017-11-08 paper/paper.tex
Add init scaling to factorization mixtures.

@@
 \section{Model}
-Let $U_i$ be a $m \times k$ matrix representing the $m$ tastes of user $i$, and $A_i$ be a $m \times k$ matrix representing the affinities of each taste for representing particular items.
+We propose three variants of the mixture-of-tasets model, covering both recurrent models and traditional factorization models.
+The \textbf{Mixture-LSTM} model uses a long-short term memory network (LSTM, \cite{hochreiter1997long}) followed by linear projection layers to obtain the mixture-of-tastes representation.
+By using a recurrent architecture, we capture both information from both the identity of the items the user interacted with as well as the order in which those interactions occurred.
+The \textbf{Projection Mixture} model .
+It is, however, sequence-oblivious, and .
+The \textbf{Embedding Mixture} model differs by directly embedding all the mixture parameters for each user.
+
+Formally, let $U_i$ be a $m \times k$ matrix representing the $m$ tastes of user $i$, and $A_i$ be a $m \times k$ matrix representing the affinities of each taste for representing particular items.
 The recommendation score for item $j$, represented by a $k times 1$-dimensional embedding vector $e_j$, is then given by
@@
 \begin{equation}
+  \label{eq:lstm}
   z_{it} = \mathrm{LSTM}\left(e_{i1}, e_{i2}, \ldots, e_{it}\right).
@@
 \end{table}
+
+\subsection{Baselines}
+Our baselines are exact equivalents of the mixture models, differing only in the fact that they represent the user with a single $k$-dimensional vector.
+For sequence-based models, we use an LSTM architecture and represent the user directly with the last hidden state of the network ($z_{it}$ from equation \ref{eq:lstm}).
+For factorization models, we use a standard latent embedding vector, corresponding directly to $z_i$ from the projection mixture model.
+This makes the baselines model particularly suitable for evaluating mixture-of-tastes representations: adding multiple tastes is the only source of any performance differences between the models.
 

### 2017-11-08 mixture/factorization.py

@@ -34,8 +34,13 @@ def hyperparameter_space():
                 'type': 'bilinear',
                 **common_space
             },
+            # {
+            #     'type': 'mixture',
+            #     **common_space
+            # },
             {
-                'type': 'mixture',
+                'type': 'mixture_init',
+                'projection_scale': hp.quniform('projection_scale', 5, 50, 5),
                 **common_space
             },
             {
@@ -71,6 +76,12 @@ def get_objective(train, validation, test):
         elif h['type'] == 'mixture':
             representation = MixtureNet(train.num_users,
                                         train.num_items,
+                                        num_components=int(h['num_components']),
+                                        embedding_dim=int(h['embedding_dim']))
+        elif h['type'] == 'mixture_init':
+            representation = MixtureNet(train.num_users,
+                                        train.num_items,
+                                        projection_scale=h['projection_scale'],
                                         num_components=int(h['num_components']),
                                         embedding_dim=int(h['embedding_dim']))
         elif h['type'] == 'nonlinear_mixture':

### 2017-11-08 mixture/factorization_representation.py

@@ -40,12 +40,14 @@ class MixtureNet(nn.Module):
     """
 
     def __init__(self, num_users, num_items, embedding_dim=32,
+                 projection_scale=1.0,
                  num_components=4):
 
         super(MixtureNet, self).__init__()
 
         self.embedding_dim = embedding_dim
         self.num_components = num_components
+        self.projection_scale = projection_scale
 
         self.user_embeddings = ScaledEmbedding(num_users, embedding_dim)
         self.item_embeddings = ScaledEmbedding(num_items, embedding_dim)
@@ -59,8 +61,8 @@ class MixtureNet(nn.Module):
                                               embedding_dim * self.num_components, bias=False)
 
         for layer in (self.taste_projection, self.attention_projection):
-            torch.nn.init.xavier_normal(layer.weight, 0.1)
-            
+            torch.nn.init.xavier_normal(layer.weight, self.projection_scale)
+
     def forward(self, user_ids, item_ids):
         """
         Compute the forward pass of the representation.

### 2017-11-08 mixture/results.py

@@ -36,6 +36,7 @@ def summarize_trials(trials):
 
     for model_type in ('pooling', 'lstm',
                        'mixture', 'mixture2',
+                       'mixture_init',
                        'linear_mixture', 'diversified_mixture',
                        'diversified_mixture_fixed',
                        'embedding_mixture',

### 2017-11-12 paper/paper.tex
Give tables consistent width.

@@
 \usepackage{caption}
+\usepackage{tabularx}
 
@@
 Formally, let $U_i$ be a $m \times k$ matrix representing the $m$ tastes of user $i$, and $A_i$ be a $m \times k$ matrix representing the affinities of each taste for representing particular items.
-The recommendation score for item $j$, represented by a $k times 1$-dimensional embedding vector $e_j$, is then given by
+The recommendation score for item $j$, represented by a $k \times 1$-dimensional embedding vector $e_j$, is then given by
 \begin{equation}
@@
 \begin{table}[htbp]
-\begin{threeparttable}
 \caption{Experimental results}
@@
 \input{includes/performance.tex}
-\begin{tablenotes}
-\small{
-\item[1] Ratio of binary model MRR to real-valued model MRR
-\item[2] Predictions Per Millisecond: how many items can be scored per millisecond
-\item[3] Ratio of binary PPMS to real-valued PPMS
-\item[4] Ratio of memory required to store binary vs. real-valued parameters
-}
-\end{tablenotes}
-\end{threeparttable}
 \end{table}

### 2017-11-12 mixture/results.py

@@ -92,11 +92,20 @@ def read_results(path, variant):
     return results
 
 
+def _full_width_table(tex):
+
+    return (tex
+            .replace('tabular', 'tabularx')
+            .replace('\\begin{tabularx}', '\\begin{tabularx}{\\columnwidth}'))
+
+
 def generate_performance_table(sequence, factorization):
 
     headers = ['Model', 'Movielens 10M', 'Amazon', 'Goodbooks']
     datasets = ('10M', 'amazon', 'goodbooks')
     rows = []
+
+    outputs = []
 
     for (model_name, model) in (('LSTM', 'lstm'),
                                 ('Mixture-LSTM', 'mixture')):
@@ -108,6 +117,15 @@ def generate_performance_table(sequence, factorization):
             row.append(mrr)
 
         rows.append(row)
+
+    outputs.append(
+        tabulate(rows,
+                 headers=headers,
+                 floatfmt='.4f',
+                 tablefmt='latex_booktabs')
+    )
+
+    rows = []
 
     for (model_name, model) in (('Bilinear', 'bilinear'),
                                 ('Projection Mixture', 'mixture'),
@@ -121,12 +139,30 @@ def generate_performance_table(sequence, factorization):
 
         rows.append(row)
 
+    outputs.append(
+        tabulate(rows,
+                 headers=headers,
+                 floatfmt='.4f',
+                 tablefmt='latex_booktabs')
+    )
+
     output = tabulate(rows,
                       headers=headers,
                       floatfmt='.4f',
                       tablefmt='latex_booktabs')
 
-    return output.replace('Bilinear', '\midrule\n Bilinear')
+    sequence_table, factorization_table = outputs
+
+    return _full_width_table(
+        '\\begin{subtable}{\\columnwidth}\n'
+        '\\caption{Sequence models}\n'
+        '%s\n'
+        '\\end{subtable}\n'
+        '\\hspace{\\fill}\n'
+        '\\begin{subtable}{\\columnwidth}\n'
+        '\\caption{Factorization models}\n'
+        '%s\n'
+        '\\end{subtable}' % (sequence_table, factorization_table))
 
 
 def generate_hyperparameter_table(sequence, factorization):
@@ -160,18 +196,20 @@ def generate_hyperparameter_table(sequence, factorization):
 
     sequence_table, factorization_table = outputs
 
-    return ('\\begin{table}\n'
[…]

### 2017-11-12 experiment/summarize_results.py
Update with newest results.

@@ -14,7 +14,7 @@ if __name__ == '__main__':
     sequence = read_results('mixture_results', 'sequence')
     factorization = read_results('mixture_results', 'factorization')
 
-    fig = plot_hyperparam_search(sequence, factorization, max_iter=25)
+    fig = plot_hyperparam_search(sequence, factorization, max_iter=20)
     fig.savefig(os.path.join(dest_dir, 'hyperparam_search.pdf'))
 
     with open(os.path.join(dest_dir, 'performance.tex'), 'w') as fle:

### 2017-11-12 paper/paper.tex
More paper writing.

@@
 
-\title{Mixture-of-interests models for representing users with diverse interests}
+\title{Mixture-of-tastes models for representing users with diverse interests}
 \author{Maciej Kula}
@@
 Most existing recommendation approaches implicitly treat user tastes as unimodal, resulting in an average-of-tastes representations when multiple distinct interests are present.
-We show that appropriately modelling the multi-faceted nature of user tastes through a mixture-of-tastes model leads to large increases in recommendation quality at a modest cost in model complexity.
+We show that appropriately modelling the multi-faceted nature of user tastes through a mixture-of-tastes model leads to large increases in recommendation quality.
 Our result holds both for deep sequence-based and traditional factorization models.
+In sequence-based models, this improvement is achieved at a very modest cost in model complexity, making mixture-of-tastes models a straightforward improvement on existing baselines.
 \end{abstract}
@@
 Latent vector-based recommender models commonly represent users with a single latent vector per user \citep{koren2009bellkor, hu2008collaborative}.
-This representation, while simple in formulation and efficient to compute, neglects te fact that users have diverse sub-tastes, and that observed user interactions can be seen as manifestations of several \emph{distinct} tastes or intents.
+This representation, while simple in formulation and efficient to compute, neglects the fact that users have diverse sub-tastes, and that observed user interactions can be seen as manifestations of several \emph{distinct} tastes or intents.
 These may either be stable facets of the user's preference (liking both horror and documentary movies, or both bluegrass and classical music), context-driven changes (preference for short-form TV content during the week but long-form cinematography during the week-end), or manifestations of phenomena like account sharing, where two (or more) different users, with correspondingly different tastes, share the same user account.
@@
 \section{Model}
-We propose three variants of the mixture-of-tasets model, covering both recurrent models and traditional factorization models.
-The \textbf{Mixture-LSTM} model uses a long-short term memory network (LSTM, \cite{hochreiter1997long}) followed by linear projection layers to obtain the mixture-of-tastes representation.
+We propose three variants of the mixture-of-tastes model, covering both recurrent models and traditional factorization models.
+
+Our sequence-based model is the Mixture LSTM (\textbf{M-LSTM}) model.
+It builds on prior work (\citet{wu2017recurrent, hidasi2015session}) using recurrent neural networks to model the sequence of user interactions.
+In our model, we use a long-short term memory network (LSTM, \cite{hochreiter1997long}) to transform the sequence of user interactions into a latent representation; we then project it into the item embedding space via a number of linear projection layers to obtain the mixture-of-tastes representation.
 By using a recurrent architecture, we capture both information from both the identity of the items the user interacted with as well as the order in which those interactions occurred.
-The \textbf{Projection Mixture} model .
-It is, however, sequence-oblivious, and .
-The \textbf{Embedding Mixture} model differs by directly embedding all the mixture parameters for each user.
+
+Within the classical matrix factorization framework, we experiment with two approaches to modelling taste mixtures.
+The first is the Projection-Mixture Factorization (\textbf{PM-F}) model.
+It extends the basic matrix factorization model by using linear projection layers on top of user latent vectors to model multiple tastes.
+In contrast, the Embedding-Mixture Factorization (\textbf{EM-F}) model directly embeds all the mixture parameters for each user.
+This makes it less parsimonious, but potenitally more expressive.
 
@@
 \end{equation}
-where $\sigma$ is the elementwise sigmoid function, $\sigma\left(A_ie_j\right)$ gives the mixture probabilities, and $U_ie_j$ the recommendation scores output by each mixture component.
+where $\sigma$ is the elementwise sigmoid function, $\sigma\left(A_ie_j\right)$ gives the mixture probabilities, and $U_ie_j$ the recommendation scores given by each mixture component.
 We assume identity variance matrices for all mixture components.
 
-This gives a mixture of tastes, where the contribution of each individual tastes component is weighted by how competent it is in evaluating item $j$.
-
 How $U_i$ and $A_i$ are obtained differs between the three evaluated models.
-In the Mixture-LSTM model, $U_{it}$ and $A_{it}$ (now indexed by $t$, their position in the sequence of interactions) are linear functions of the hidden state of an LSTM layer trained on the user's previous interactions, $z_{it}$:
+In the M-LSTM model, $U_{it}$ and $A_{it}$ (now indexed by $t$, their position in the sequence of interactions) are linear functions of the hidden state $z_{it}$ of an LSTM layer trained on the user's previous interactions:
 \begin{equation}
@@
 \end{equation}
-Given $z_{it}$, the $m$-th row of $U_{it}$ and $A_{it}$ is given by
+Given $z_{it}$, the $m$-th row of $U_{it}$ and $A_{it}$ are given by
 \begin{equation}
@@
 
-The Projection Mixture factorization model is similar: an embedding $z_i$ is estimated for each user, and the $U_i$ and $A_i$ matrices are obtained via linear projections:
+The PM-F model is similar: an embedding $z_i$ is estimated for each user, and the $U_i$ and $A_i$ matrices are obtained via linear projections:
 \begin{equation}
@@
 This keeps the number of model parameters small at a potential cost to model capacity.
-In contrast, the Embedding Mixture factorization model embeds $U_i$ and $A_i$ directly.
+In contrast, the EM-F model embeds $U_i$ and $A_i$ directly.
 This substantially increases the number of model parameters, but may lead to better accuracy.
 
-In all models, the input and output item embeddings $e$ are tied.
+In all models, the item representations are given by the latent vectors $e$. The input and output item embeddings are tied, and they have the same dimensionality as the user representations.
 
 \section{Experiments}
+We test our models on a number of publicly available datasets with varying degrees of sparsity as well as diversity of tastes.
+In all tests, we treat our models as ranking models, and evaluate them on the quality of the ranked recommendation list they generate, rather than the rating predictions they produce (unlike \citep{wu2017recurrent}).
+
+% A priori, we expect the mixture models to perform at least as well as non-mixture models, given that
+
 \subsection{Datasets}
-We use the following datasets in out experiments.
+We use the following datasets in our experiments (their summary statistics are listed in Table \ref{tab:datasets}):
 
@@
 \end{enumerate}
+The two key differences betweeen the datasets are their sparsity and the degree to which they are popularity-biased.
+The Movielens 10M and Goodbooks datasets are relatively dense, while the Amazon dataset is much sparser.
+Popularity seems to play a greater role in the Movielens dataset than in the other datasets: the ratio between the number of interactions that accrue to the 95th percentile and the 50th percentile of most popular items is highest in Movielens 10M.
 
-We treat all datasets as implicit feedback datasets, where the existence of an edge between a user and an item expresses implicit preference, and the lack of an edge implicit lack of preference.
+We conjecture that taste diversity plays a lesser role in highly popularity-biased datasets (that is, where a large share of all interactions go to very few items).
+Intuitively, a single dominant taste can model such observations quite well; additional tastes only come into play in the long tail of the distribution.
+Conversely, a more even distribution of interactions between items allows the possibility that multiple tastes play a larger role.
+If this is true, we would expect the gains from our mixture models to be larger in the Amazon (where tastes can span multiple unrelated item categories) and Goodbooks (where the popularity distribution is relatively even by construction) datasets than in the Movielens dataset.
+
+Throughout our experiments, we treat all datasets as implicit feedback datasets, where the existence of an edge between a user and an item expresses implicit preference, and the lack of an edge implicit lack of preference.
 
 \begin{table}
-  \caption{Dataset statistics}
+  \caption{Dataset statistics. $95th/50th$ denotes the ratio of popularity of the 95th and 50th percentile of item popularity.}
   \label{tab:datasets}
@@
 For factorization models, we use a standard latent embedding vector, corresponding directly to $z_i$ from the projection mixture model.
-This makes the baselines model particularly suitable for evaluating mixture-of-tastes representations: adding multiple tastes is the only source of any performance differences between the models.
+
+This makes our baselines particularly suitable for evaluating mixture-of-tastes representations: adding multiple tastes is the sole architectural difference between the models, and so differences in recommendation performance can be directly attributed to the models' greater ability to model diverse tastes.
 
@@
 
+We believe our sequence model experimental setting to be a relatively good reflection of the conditions in which industry recommender system are trained and evaluated: model retraining is often done daily, with data up to the day of training used for model estimation, and subsequent interactions used for evaluation.
+Insofar as this is true, our results should generalize well to real-world applications.
+
+\subsection{Loss functions}
+
 We experiment with two loss functions:
@@
 \end{itemize}
-For both loss functions, for any known positive user-item interaction pair $(i, j)$, we uniformly sample an implicit negative item $k \in \symUserInteractionSet^-$.
-For BPR, the loss for any such triplet is given by
+For both loss functions, for any known positive user-item interaction pair $(i, j)$, we uniformly sample an implicit negative item $k$. For BPR, the loss for any such triplet is given by
 \begin{equation}
@@
 
+\subsection{Hyperparameter search}
+
 We perform extensive hyperparameter optimization across both our proposed models and all baselines.
@@
 
+\section{Implementation}
+Our models are implemented using the PyTorch deep learning framework \citep{paszke2017pytorch} and trained using the nVidia K40 GPUs.
+We make the model and experiment code (as well as the full results) available on Github\footnote{\url{https://github.com/maciejkula/mixture}}.
+
 \section{Results}
+\begin{table}[htbp]
+\caption{Experimental results}
+\label{tb:results}
+\input{includes/performance.tex}
+\end{table}
+Our results are summarized in Table \ref{tb:results}.
 
@@
 By and large, there is a dose-response relationship between the number of mixtures and recommendation quality: being able to represents more distinct user tasts yields better results.
-
-\begin{table}[htbp]
-\caption{Experimental results}
-\label{tb:results}
-\input{includes/performance.tex}
-\end{table}
 

### 2017-11-12 mixture/results.py

@@ -270,7 +270,7 @@ def plot_hyperparam_search(sequence, factorization, max_iter=100):
 
 def generate_dataset_table():
 
-    headers = ['Dataset', 'Users', 'Items', 'Interactions', 'Density']
+    headers = ['Dataset', 'Users', 'Items', 'Density', '95th/50th']
 
     rows = []
 
@@ -278,12 +278,17 @@ def generate_dataset_table():
                           ('Amazon', get_amazon_dataset()),
                           ('Goodbooks', get_goodbooks_dataset())):
 
+        item_counts = dataset.tocoo().getnnz(axis=1)
+
+        print('Dataset {}, ratio: {:0,}'
+              .format(name, np.percentile(item_counts, 95) / np.percentile(item_counts, 50)))
+
         row = [
             name,
             '{:0,}'.format(dataset.num_users),
             '{:0,}'.format(dataset.num_items),
-            '{:0,}'.format(len(dataset)),
-            len(dataset) / dataset.num_users / dataset.num_items
+            len(dataset) / dataset.num_users / dataset.num_items,
+            '{0:.2f}'.format(np.percentile(item_counts, 95) / np.percentile(item_counts, 50))
         ]
 
         rows.append(row)

### 2017-11-12 paper/paper.tex
Results section.

@@
-\documentclass[sigconf]{acmart}
+\documentclass[sigchi]{acmart}
 
@@
 We show that appropriately modelling the multi-faceted nature of user tastes through a mixture-of-tastes model leads to large increases in recommendation quality.
-Our result holds both for deep sequence-based and traditional factorization models.
+Our result holds both for deep sequence-based and traditional factorization models, and is robust to careful selection and tuning of baseline models.
 In sequence-based models, this improvement is achieved at a very modest cost in model complexity, making mixture-of-tastes models a straightforward improvement on existing baselines.
@@
 
+Throughout, we take a take great care to select and carefully tune our baseline models.
+
 \section{Related work}
@@
 Secondly, large (and non-spherical) variances aid the representation of polysemous words.
-In the context of recommendations, Gausian embeddings with large variances could be used to represent users with wide-ranging tastes.
+In the context of recommendations, Gaussian embeddings with large variances could be used to represent users with wide-ranging tastes.
 
@@
 \end{enumerate}
+Throughout our experiments, we treat all datasets as implicit feedback datasets, where the existence of an edge between a user and an item expresses implicit preference, and the lack of an edge implicit lack of preference.
+
 The two key differences betweeen the datasets are their sparsity and the degree to which they are popularity-biased.
@@
 If this is true, we would expect the gains from our mixture models to be larger in the Amazon (where tastes can span multiple unrelated item categories) and Goodbooks (where the popularity distribution is relatively even by construction) datasets than in the Movielens dataset.
-
-Throughout our experiments, we treat all datasets as implicit feedback datasets, where the existence of an edge between a user and an item expresses implicit preference, and the lack of an edge implicit lack of preference.
 
@@
 
-\section{Implementation}
+\subsection{Implementation}
 Our models are implemented using the PyTorch deep learning framework \citep{paszke2017pytorch} and trained using the nVidia K40 GPUs.
+All of the models are trained using the ADAM \citep{kingma2014adam} per-parameter learning rate schedule.
 We make the model and experiment code (as well as the full results) available on Github\footnote{\url{https://github.com/maciejkula/mixture}}.
@@
 \begin{table}[htbp]
-\caption{Experimental results}
+\caption{Mean reciprocal rank across all users/sequences in the test set.
+Note that due to differences in experimental protocol, results between sequence-based and factorization models are not directly comparable.}
 \label{tb:results}
@@
 \end{table}
-Our results are summarized in Table \ref{tb:results}.
+Our main results are summarized in Table \ref{tb:results}.
+In both sequence-based and factorization tasks variants of our model achieve consistently higher performance than baseline models, with improvements ranging from 10\% to 69\%.
+The results are robust to optimizing the hyperparameters of our baselines, giving us ample confidence that the results can be replicated in production recommender systems.
+
+\subsection{Ranking quality}
+In sequence-based models, our M-LSTM model outperforms the baseline LSTM model on all datasets.
+The performance gains are particularly large on the Amazon (26\%) datset and Goodbooks (17\%) datasets, and smaller, but still meaningful, on the Movielens 10M dataset (10\%).
+This is consistent with our conjectures on the nature of the datasets.
+The Amazon dataset, spanning the entire catalog of Amazon products, benefits most from being able to model taste diversity.
+This is not surprising, given that a single Amazon user can interact with products from many disparate domains.
+
+Among factorization models, the Embedding Mixture model is the best performing model in two datasets.
+It beats the baseline model by a very substantial margin on the Amazon dataset (69\%), and by a smaller margin of 12\% on the Goodbooks dataset.
+There is relatively little to distinguish the models on the Movielens dataset, where all three models perform equally well.
+Somewhat surprisngly, the Projection Mixture model fails to outperform the baseline on two datasets, and substantially \emph{underperforms} it on the Amazon dataset.
+Given that the model is strictly more expressive than the baseline model, we attribute this failure to difficulties in optimization that have eluded us despite experimenting with a number of initialization schemes and learning rate schedules.
 
 \subsection{Hyperparameter search}
+\begin{figure*}[h]
+  \captionsetup{width=.8\linewidth}
+  \caption{Maximum test MRR vs number of hyperparameter search iterations. Sequence-based models in the top row; factorization-based models in the bottom row.}
+  \label{fig:hyper}
+  \includegraphics[width=0.8\textwidth]{includes/hyperparam_search.pdf}
+\end{figure*}
+Our results are robust to hyperparameter optimization.
 Figure~\ref{fig:hyper} plots the maximum test MRR achieved by each algorithm as a function of the number of elapsed hyperparameter search iterations.
@@
 \subsection{Number of taste components}
+\input{includes/hyperparameters.tex}
 We summarize the effect of increasing the number of taste mixture components in Table~\ref{tab:nummixtures}.
 By and large, there is a dose-response relationship between the number of mixtures and recommendation quality: being able to represents more distinct user tasts yields better results.
+The optimum number of components is dataset dependent: the (more diverse) Amazon and Goodbooks dataset tend to benefit more from additional expressiveness, while four mixture components seem to suffice in the more homogeneous Movielens dataset.
 
-\input{includes/hyperparameters.tex}
-
-\begin{figure*}[h]
-  \captionsetup{width=.8\linewidth}
-  \caption{Maximum test MRR vs number of hyperparameter search iterations. Sequence-based models in the top row; factorization-based models in the bottom row.}
-  \label{fig:hyper}
-  \includegraphics[width=0.8\textwidth]{includes/hyperparam_search.pdf}
-\end{figure*}
+\section{Conclusion}
 

### 2017-11-12 paper/paper.tex
Add conclusion.

@@
 \section{Conclusion}
+We show that mixture-of-tastes representations are clear improvements over their baseline modes, especially in sequence-based settings where the accuracy gains come at a very modest cost in model complexity.
+We have taken care to test the mixture models against strong baselines, and have confidence that our approach can translate to accuracy gains in production recommender systems.
 

### 2017-11-12 paper/paper.tex
Change title.
Make changes to use AAAI format.

@@
-\documentclass[sigchi]{acmart}
+\documentclass[letterpaper]{article}
+\usepackage{aaai}
+\usepackage{times}
+\usepackage{helvet}
+\usepackage{courier}
+\usepackage{url}
+\usepackage{graphicx}
+\frenchspacing
+% Add additional packages here. The following
+% packages may NOT be used (this list
+% is not exhaustive:
+% authblk, caption, CJK, float, fullpage, geometry,
+%hyperref, layout, nameref, natbib, savetrees,
+%setspace, titlesec, tocbibind, ulem
+%
+
+%
+%
+% PDFINFO
+% You are required to complete the following
+% for pass-through to the PDF.
+% No LaTeX commands of any kind may be
+% entered. The parentheses and spaces
+% are an integral part of the
+% pdfinfo script and must not be removed.
+%
+\pdfinfo{
+/Title (Mixture-of-tastes Models for Representing Users with Diverse Interests)
+/Author (Maciej Kula)
+/Keywords (Recommender Systems, Matrix Factorization, Sequence-based Recommendations)
+}
+%
+%Section Numbers
+% Uncomment if you want to use section numbers
+% and change the 0 to a 1 or 2
+% \setcounter{secnumdepth}{0}
+% Title and Author Information Must Immediately Follow
+% the pdfinfo within the preamble
+%
 
@@
 \usepackage{tabularx}
+\usepackage{natbib}
 
@@
 
-\title{Mixture-of-tastes models for representing users with diverse interests}
-\author{Maciej Kula}
-\email{maciej.kula@gmail.com}
+%US Lettersize Paper Is Required
+
+\title{Mixture-of-tastes Models for Representing Users with Diverse Interests}
+\author{Maciej Kula\\maciej.kula@gmail.com}
+%\email{maciej.kula@gmail.com}
 \date{\today}
-\settopmatter{printacmref=false, printccs=false, printfolios=true}
-\setcopyright{none}
-\acmConference[]{}{}{}
+%\settopmatter{printacmref=false, printccs=false, printfolios=true}
+%\setcopyright{none}
+%\acmConference[]{}{}{}
 
-\begin{CCSXML}
-<ccs2012>
-<concept>
-<concept_id>10002951.10003260.10003261.10003269</concept_id>
-<concept_desc>Information systems~Collaborative filtering</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-<concept>
-<concept_id>10002951.10003317.10003338.10003346</concept_id>
-<concept_desc>Information systems~Top-k retrieval in databases</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-</ccs2012>
-\end{CCSXML}
+%% \begin{CCSXML}
+%% <ccs2012>
+%% <concept>
+%% <concept_id>10002951.10003260.10003261.10003269</concept_id>
+%% <concept_desc>Information systems~Collaborative filtering</concept_desc>
+%% <concept_significance>500</concept_significance>
+%% </concept>
+%% <concept>
+%% <concept_id>10002951.10003317.10003338.10003346</concept_id>
+%% <concept_desc>Information systems~Top-k retrieval in databases</concept_desc>
+%% <concept_significance>500</concept_significance>
+%% </concept>
+%% </ccs2012>
+%% \end{CCSXML}
 
-\ccsdesc[500]{Information systems~Collaborative filtering}
-\ccsdesc[500]{Information systems~Top-k retrieval in databases}
+%\ccsdesc[500]{Information systems~Collaborative filtering}
+% \ccsdesc[500]{Information systems~Top-k retrieval in databases}
 
-\keywords{Recommender Systems, Matrix Factorization, Sequence-based Recommendations}
+% \keywords{Recommender Systems, Matrix Factorization, Sequence-based Recommendations}
+
+\maketitle
 
@@
 \end{abstract}
-
-\maketitle
 
@@
   \begin{figure*}[h]
+    \centering
     \captionsetup{width=.8\linewidth}
@@
   \end{figure*}
+
 Our results are robust to hyperparameter optimization.
@@
 \bibliography{bibliography}
-\bibliographystyle{ACM-Reference-Format}
+\bibliographystyle{aaai}
 

### 2017-11-13 paper/paper.tex

@@
 Latent vector-based recommender models commonly represent users with a single latent vector per user \citep{koren2009bellkor, hu2008collaborative}.
-This representation, while simple in formulation and efficient to compute, neglects the fact that users have diverse sub-tastes, and that observed user interactions can be seen as manifestations of several \emph{distinct} tastes or intents.
-These may either be stable facets of the user's preference (liking both horror and documentary movies, or both bluegrass and classical music), context-driven changes (preference for short-form TV content during the week but long-form cinematography during the week-end), or manifestations of phenomena like account sharing, where two (or more) different users, with correspondingly different tastes, share the same user account.
+This representation, while simple and efficient to compute, neglects the fact that users have diverse sub-tastes, and that observed user interactions can be seen as manifestations of several distinct tastes or intents.
+These may either be stable facets of the user's preference (liking both horror and documentary movies, or both bluegrass and classical music), context-driven changes (preference for short-form TV content during the week but long-form cinematography during the week-end), or manifestations of phenomena like account sharing, where two different users, with correspondingly different tastes, share the same user account.
 
@@
 In the case of deep recurrent sequence models, this improvement is achieved at a very modest cost to to the model complexity, making our model a straightforward improvement over existing methods.
-
-Throughout, we take a take great care to select and carefully tune our baseline models.
 
@@
 \section{Model}
-We propose three variants of the mixture-of-tastes model, covering both recurrent models and traditional factorization models.
+We propose three variants of the mixture-of-tastes model, covering both recurrent and traditional factorization models.
 
 Our sequence-based model is the Mixture LSTM (\textbf{M-LSTM}) model.
-It builds on prior work (\citet{wu2017recurrent, hidasi2015session}) using recurrent neural networks to model the sequence of user interactions.
+It builds on prior work \citep{wu2017recurrent, hidasi2015session} using recurrent neural networks to model the sequence of user interactions.
 In our model, we use a long-short term memory network (LSTM, \cite{hochreiter1997long}) to transform the sequence of user interactions into a latent representation; we then project it into the item embedding space via a number of linear projection layers to obtain the mixture-of-tastes representation.
-By using a recurrent architecture, we capture both information from both the identity of the items the user interacted with as well as the order in which those interactions occurred.
+By using a recurrent architecture, we capture information from the identity of the items the user interacted with as well as the order in which those interactions occurred.
 
@@
 In contrast, the Embedding-Mixture Factorization (\textbf{EM-F}) model directly embeds all the mixture parameters for each user.
-This makes it less parsimonious, but potenitally more expressive.
+This makes it less parsimonious, but potentially more expressive.
 
@@
 \begin{equation}
-  r_{ij} = \sigma\left(A_ie_j\right) \cdot U_ie_j,
+  r_{ij} = \sigma\left(A_ie_j\right) \cdot U_ie_j + b_j,
 \end{equation}
-where $\sigma$ is the elementwise sigmoid function, $\sigma\left(A_ie_j\right)$ gives the mixture probabilities, and $U_ie_j$ the recommendation scores given by each mixture component.
+where $b_j$ is the per-item bias term, $\sigma$ the elementwise sigmoid function, $\sigma\left(A_ie_j\right)$ gives the mixture probabilities, and $U_ie_j$ the recommendation scores given by each mixture component.
 We assume identity variance matrices for all mixture components.
@@
 \end{equation}
-where $W^U_m$, $W^A_M$, $B^U_m$, and $B^A_m$ are the learned projection matrices.
-These are common across all users, representing a modest increase in the total number of model parameters.
+where $W^U_m$, $W^A_M$ are the learned projection matrices, and $B^U_m$ and $B^A_m$ are bias terms.
+Both the projection matrices and the biases are common across all users, representing a modest increase in the total number of model parameters.
 Note that the LSTM network only needs to be run once to obtain the full user representation.
@@
 We test our models on a number of publicly available datasets with varying degrees of sparsity as well as diversity of tastes.
-In all tests, we treat our models as ranking models, and evaluate them on the quality of the ranked recommendation list they generate, rather than the rating predictions they produce (unlike \citep{wu2017recurrent}).
+In all tests, we treat our models as ranking models, and evaluate them on the quality of the ranked recommendation list they generate, rather than the rating predictions they produce (unlike \cite{wu2017recurrent}).
 
@@
 \item Goodbooks: dataset of 6 million ratings across 53,000 users and 10,000 most popular books from the Goodbooks online book recommendation and sharing service \citep{goodbooks2017}.
-\item Amazon: dataset of ratings and reviews gathered from the Amazon online shopping service \citep{leskovec2007dynamics}.
+\item Amazon: dataset of ratings and reviews gathered from the Amazon online shopping service, spanning books, music, and videos \citep{leskovec2007dynamics}.
 After pruning users and items with fewer than 10 ratings, the dataset contains approximately 4 million ratings from 100,000 users over 114,000 items.
 \end{enumerate}
-Throughout our experiments, we treat all datasets as implicit feedback datasets, where the existence of an edge between a user and an item expresses implicit preference, and the lack of an edge implicit lack of preference.
-
 The two key differences betweeen the datasets are their sparsity and the degree to which they are popularity-biased.
@@
 
+Throughout our experiments, we treat all datasets as implicit feedback datasets, where the existence of an edge between a user and an item expresses implicit preference, and the lack of an edge implicit lack of preference.
+
 \begin{table}
-  \caption{Dataset statistics. $95th/50th$ denotes the ratio of popularity of the 95th and 50th percentile of item popularity.}
+  \caption{Dataset statistics. 95th/50th denotes the ratio of popularity of the 95th and 50th percentile of item popularity.}
   \label{tab:datasets}
@@
 \subsection{Loss functions}
-
 We experiment with two loss functions:
@@
 Our goal is two-fold.
-Firstly, we want to mitigate researcher bias, where more care and attention is devoted to the researcher's proposed model, thus unfairly disadvantaging baseline algorithms in a performance comparison.
-We believe this to be a common phenomenon; its extent is illustrated, in a related domain, by \cite{melis2017state}, who find that standard LSTM architectures, when properly tuned, outperform more recent algorithms in natural laguage modelling tasks.
+Firstly, we want to mitigate researcher bias, where more care and attention is devoted to the researcher's proposed model, thus unfairly disadvantaging baseline algorithms.
+We believe this to be a common phenomenon; its extent is illustrated, in a related domain, by \cite{melis2017state}, who find that standard LSTM architectures when properly tuned outperform more recent algorithms in natural laguage modelling tasks.
 Secondly, we wish to understand the extent to which the mixture-of-interests models are \emph{fragile}, in the sense of being highly sensitive to hyperparameter choices.
@@
 \section{Results}
+Our main results are summarized in Table \ref{tb:results}.
+In both sequence-based and factorization tasks, variants of our model achieve consistently higher performance than baseline models, with improvements ranging from 10\% to 69\%.
+The results are robust to optimizing the hyperparameters of our baselines, giving us ample confidence that the results can be replicated in production recommender systems.
 \begin{table}[htbp]
@@
 \end{table}
-Our main results are summarized in Table \ref{tb:results}.
-In both sequence-based and factorization tasks variants of our model achieve consistently higher performance than baseline models, with improvements ranging from 10\% to 69\%.
-The results are robust to optimizing the hyperparameters of our baselines, giving us ample confidence that the results can be replicated in production recommender systems.
 
@@
 Somewhat surprisngly, the Projection Mixture model fails to outperform the baseline on two datasets, and substantially \emph{underperforms} it on the Amazon dataset.
-Given that the model is strictly more expressive than the baseline model, we attribute this failure to difficulties in optimization that have eluded us despite experimenting with a number of initialization schemes and learning rate schedules.
+Given that the model is strictly more expressive than the baseline model, we attribute this failure to difficulties in effectively fitting the model.
 

### 2017-11-14 paper/paper.tex
Factual and spelling fixes.

@@
-\documentclass[letterpaper]{article}
-\usepackage{aaai}
-\usepackage{times}
-\usepackage{helvet}
-\usepackage{courier}
-\usepackage{url}
-\usepackage{graphicx}
-\frenchspacing
-% Add additional packages here. The following
-% packages may NOT be used (this list
-% is not exhaustive:
-% authblk, caption, CJK, float, fullpage, geometry,
-%hyperref, layout, nameref, natbib, savetrees,
-%setspace, titlesec, tocbibind, ulem
-%
-
-%
-%
-% PDFINFO
-% You are required to complete the following
-% for pass-through to the PDF.
-% No LaTeX commands of any kind may be
-% entered. The parentheses and spaces
-% are an integral part of the
-% pdfinfo script and must not be removed.
-%
-\pdfinfo{
-/Title (Mixture-of-tastes Models for Representing Users with Diverse Interests)
-/Author (Maciej Kula)
-/Keywords (Recommender Systems, Matrix Factorization, Sequence-based Recommendations)
-}
-%
-%Section Numbers
-% Uncomment if you want to use section numbers
-% and change the 0 to a 1 or 2
-% \setcounter{secnumdepth}{0}
-% Title and Author Information Must Immediately Follow
-% the pdfinfo within the preamble
-%
+\documentclass[sigconf]{acmart}
 
@@
 \title{Mixture-of-tastes Models for Representing Users with Diverse Interests}
-\author{Maciej Kula\\maciej.kula@gmail.com}
-%\email{maciej.kula@gmail.com}
+\author{Maciej Kula}
+\email{maciej.kula@gmail.com}
 \date{\today}
-%\settopmatter{printacmref=false, printccs=false, printfolios=true}
-%\setcopyright{none}
-%\acmConference[]{}{}{}
+\settopmatter{printacmref=false, printccs=false, printfolios=true}
+\setcopyright{none}
+\acmConference[]{}{}{}
 
-%% \begin{CCSXML}
-%% <ccs2012>
-%% <concept>
-%% <concept_id>10002951.10003260.10003261.10003269</concept_id>
-%% <concept_desc>Information systems~Collaborative filtering</concept_desc>
-%% <concept_significance>500</concept_significance>
-%% </concept>
-%% <concept>
-%% <concept_id>10002951.10003317.10003338.10003346</concept_id>
-%% <concept_desc>Information systems~Top-k retrieval in databases</concept_desc>
-%% <concept_significance>500</concept_significance>
-%% </concept>
-%% </ccs2012>
-%% \end{CCSXML}
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10002951.10003260.10003261.10003269</concept_id>
+<concept_desc>Information systems~Collaborative filtering</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10002951.10003317.10003338.10003346</concept_id>
+<concept_desc>Information systems~Top-k retrieval in databases</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
 
-%\ccsdesc[500]{Information systems~Collaborative filtering}
-% \ccsdesc[500]{Information systems~Top-k retrieval in databases}
+\ccsdesc[500]{Information systems~Collaborative filtering}
+\ccsdesc[500]{Information systems~Top-k retrieval in databases}
 
-% \keywords{Recommender Systems, Matrix Factorization, Sequence-based Recommendations}
-
-\maketitle
+%\keywords{Recommender Systems, Matrix Factorization, Sequence-based Recommendations}
 
@@
 
+\maketitle
+
 \section{Introduction}
@@
 This representation, while simple and efficient to compute, neglects the fact that users have diverse sub-tastes, and that observed user interactions can be seen as manifestations of several distinct tastes or intents.
-These may either be stable facets of the user's preference (liking both horror and documentary movies, or both bluegrass and classical music), context-driven changes (preference for short-form TV content during the week but long-form cinematography during the week-end), or manifestations of phenomena like account sharing, where two different users, with correspondingly different tastes, share the same user account.
+These may either be stable facets of the user's preference (liking both horror and documentary movies, or both bluegrass and classical music), context-driven changes (preference for short-form TV content during the week but long-form cinematography during the weekend), or manifestations of phenomena like account sharing, where two different users, with correspondingly different tastes, share the same user account.
 
@@
 Firstly, it may lead to lack of nuance in representing the user, where a dominant taste may overpower more niche ones.
-Secondly it may reduce the quality of item representations, especially when it leads to decreasing the separation in the embedding space between groups of items belonging to multiple tastes or genres.
-For illustration, suppose that documentaries and horror movies are distinct genres, in the sense that most users prefer one or the other: but that there is still a group of users who like both.
+Secondly, it may reduce the quality of item representations, especially when it leads to decreasing the separation in the embedding space between groups of items belonging to multiple tastes or genres.
+For illustration, documentaries and horror movies are distinct genres in the sense that most users prefer one or the other, but there are still users who like both.
 To represent that group of users, the genres' embeddings will have to be separated less cleanly than they would be if the model could express the concept of multiple tastes.
@@
 We apply this model to both traditional implicit feedback factorization models, and to more recent models using recurrent neural networks.
-In both cases, our mixture-of-interests models handily outperform single-representation models on standard ranking quality metrics.
+In both cases, our mixture-of-interests models outperform single-representation models on standard ranking quality metrics.
 In the case of deep recurrent sequence models, this improvement is achieved at a very modest cost to to the model complexity, making our model a straightforward improvement over existing methods.
@@
 
-How $U_i$ and $A_i$ are obtained differs between the three evaluated models.
+How $U_i$ and $A_i$ are obtained differs betwen the three evaluated models.
 In the M-LSTM model, $U_{it}$ and $A_{it}$ (now indexed by $t$, their position in the sequence of interactions) are linear functions of the hidden state $z_{it}$ of an LSTM layer trained on the user's previous interactions:
@@
 
-We believe our sequence model experimental setting to be a relatively good reflection of the conditions in which industry recommender system are trained and evaluated: model retraining is often done daily, with data up to the day of training used for model estimation, and subsequent interactions used for evaluation.
+We believe our sequence model experimental setting to be a relatively good reflection of the conditions in which industry recommender systems are trained and evaluated: model retraining is often done daily, with data up to the day of training used for model estimation, and subsequent interactions used for evaluation.
 Insofar as this is true, our results should generalize well to real-world applications.
@@
 Secondly, we wish to understand the extent to which the mixture-of-interests models are \emph{fragile}, in the sense of being highly sensitive to hyperparameter choices.
-Such fragile algorithms are potentially of lesser utility in industry applications, where the engineering cost of tuning in maintaing them may outweigh the accuracy benefits they bring.
+Such fragile algorithms are potentially of lesser utility in industry applications, where the engineering cost of tuning and maintaing them may outweigh the accuracy benefits they bring.
 
-We use random search to tune the algorthms used in our experiments.
+We use random search to tune the algorithms used in our experiments.
 We optimize batch size, number of training epochs, the learning rate, L2 regularization weight, the loss function, and (where appropriate) the number of taste mixture components.
@@
 The results are robust to optimizing the hyperparameters of our baselines, giving us ample confidence that the results can be replicated in production recommender systems.
-\begin{table}[htbp]
+\begin{table}
 \caption{Mean reciprocal rank across all users/sequences in the test set.
@@
 \subsection{Ranking quality}
+  \begin{figure*}[h]
+    \centering
+    \captionsetup{width=.8\linewidth}
+    \caption{Maximum test MRR vs number of hyperparameter search iterations. Sequence-based models in the top row; factorization-based models in the bottom row.}
+    \label{fig:hyper}
+    \includegraphics[width=0.8\textwidth]{includes/hyperparam_search.pdf}
+  \end{figure*}
 In sequence-based models, our M-LSTM model outperforms the baseline LSTM model on all datasets.
@@
 There is relatively little to distinguish the models on the Movielens dataset, where all three models perform equally well.
-Somewhat surprisngly, the Projection Mixture model fails to outperform the baseline on two datasets, and substantially \emph{underperforms} it on the Amazon dataset.
+Somewhat surprisingly, the Projection Mixture model fails to outperform the baseline on two datasets, and substantially \emph{underperforms} it on the Amazon dataset.
 Given that the model is strictly more expressive than the baseline model, we attribute this failure to difficulties in effectively fitting the model.
@@
 \subsection{Hyperparameter search}
-  \begin{figure*}[h]
-    \centering
-    \captionsetup{width=.8\linewidth}
-    \caption{Maximum test MRR vs number of hyperparameter search iterations. Sequence-based models in the top row; factorization-based models in the bottom row.}
-    \label{fig:hyper}
-    \includegraphics[width=0.8\textwidth]{includes/hyperparam_search.pdf}
-  \end{figure*}
-
 Our results are robust to hyperparameter optimization.
@@
 All algorithms converge to their optimum performance relatively quickly, suggesting a degree of robustness to hyperparameter choices.
-Mixture-LSTM and Embedding Mixture models quickly outperform their baseline counterparts, and maintain a stable performance lead thereafter.
+Mixture-LSTM and Embedding Mixture models quickly outperform their baseline counterparts, and maintain a stable performance lead thereafter (with the exception of the factorization Movielens experiments).
 This lends support to our belief that the mixture models' superior accuracy reflects their greater capacity to model the recommendation problem well, rather than being an artifact of the experimental procedure or researcher bias.
@@
 We summarize the effect of increasing the number of taste mixture components in Table~\ref{tab:nummixtures}.
-By and large, there is a dose-response relationship between the number of mixtures and recommendation quality: being able to represents more distinct user tasts yields better results.
-The optimum number of components is dataset dependent: the (more diverse) Amazon and Goodbooks dataset tend to benefit more from additional expressiveness, while four mixture components seem to suffice in the more homogeneous Movielens dataset.
+The optimum number of components is dataset and algorithm dependent but, by and large, there is a dose-response relationship between the number of mixtures and recommendation quality: being able to represent more distinct user tastes yields better results.
 
@@
 \bibliography{bibliography}
-\bibliographystyle{aaai}
+\bibliographystyle{ACM-Reference-Format}
 

### 2017-11-16 paper/paper.tex
More spelling fixes.

@@
 
-How $U_i$ and $A_i$ are obtained differs betwen the three evaluated models.
+How $U_i$ and $A_i$ are obtained differs between the three evaluated models.
 In the M-LSTM model, $U_{it}$ and $A_{it}$ (now indexed by $t$, their position in the sequence of interactions) are linear functions of the hidden state $z_{it}$ of an LSTM layer trained on the user's previous interactions:
@@
 \end{enumerate}
-The two key differences betweeen the datasets are their sparsity and the degree to which they are popularity-biased.
+The two key differences between the datasets are their sparsity and the degree to which they are popularity-biased.
 The Movielens 10M and Goodbooks datasets are relatively dense, while the Amazon dataset is much sparser.

### 2017-11-19 paper/paper.tex
Fix references to goodbooks-10k.

@@
 \item Movielens 10M: dataset of 10 million movie ratings across 10,000 movies and 72,000 users \citep{harper2016movielens}.
-\item Goodbooks: dataset of 6 million ratings across 53,000 users and 10,000 most popular books from the Goodbooks online book recommendation and sharing service \citep{goodbooks2017}.
+\item Goodbooks-10K: dataset of 6 million ratings across 53,000 users and 10,000 most popular books from the Goodreads online book recommendation and sharing service \citep{goodbooks2017}.
 \item Amazon: dataset of ratings and reviews gathered from the Amazon online shopping service, spanning books, music, and videos \citep{leskovec2007dynamics}.
@@
 We have taken care to test the mixture models against strong baselines, and have confidence that our approach can translate to accuracy gains in production recommender systems.
+We believe that modelling the diversity of user tastes in this manner can become a standard component of recommendation algorithms.
 

### 2017-12-13 paper/paper.tex
Inline results.

@@
   \label{tab:datasets}
-  \input{includes/datasets.tex}
+\begin{tabularx}{\columnwidth}{lllrr}
+\toprule
+ Dataset       & Users   & Items   &   Density &   95th/50th \\
+\midrule
+ Movielens & 69,879  & 10,678  &    0.0134 &      7.42 \\
+ Amazon        & 100,085 & 113,997 &    0.0003 &      5.67 \\
+ Goodbooks     & 53,425  & 10,001  &    0.0112 &      1.41 \\
+\bottomrule
+\end{tabularx}
 \end{table}
@@
 \label{tb:results}
-\input{includes/performance.tex}
+\begin{subtable}{\columnwidth}
+\caption{Sequence models}
+\begin{tabularx}{\columnwidth}{lrrr}
+\toprule
+ Model        &   Movielens &   Amazon &   Goodbooks \\
+\midrule
+ LSTM         &          0.0908 &   0.1502 &      0.1158 \\
+ Mixture-LSTM &          \textbf{0.1001} &   \textbf{0.1889} &      \textbf{0.1358} \\
+\bottomrule
+\end{tabularx}
+\end{subtable}
+\hspace{\fill}
+\begin{subtable}{\columnwidth}
+\caption{Factorization models}
+\begin{tabularx}{\columnwidth}{lrrr}
+\toprule
+ Model              &   Movielens &   Amazon &   Goodbooks \\
+\midrule
+ Factorization           &          0.1053 &   0.1001 &      0.0738 \\
+ Projection Mixture &          \textbf{0.1097} &   0.0698 &      0.0712 \\
+ Embedding Mixture  &          0.1026 &   \textbf{0.1696} &      \textbf{0.0853} \\
+\bottomrule
+\end{tabularx}
+\end{subtable}
+
 \end{table}
@@
     \label{fig:hyper}
-    \includegraphics[width=0.8\textwidth]{includes/hyperparam_search.pdf}
+    \includegraphics[width=0.8\textwidth]{hyperparam_search.pdf}
   \end{figure*}
@@
 \subsection{Number of taste components}
-\input{includes/hyperparameters.tex}
+\begin{table}
+\caption{Effect of number of mixture components}
+\label{tab:nummixtures}
+\begin{subtable}{\columnwidth}
+\caption{Sequence models}
+\begin{tabularx}{\columnwidth}{rrrr}
+\toprule
+   Components &   Movielens 10M &   Amazon &   Goodbooks \\
+\midrule
+            2 &          0.0882 &   0.1538 &      0.1262 \\
+            4 &          0.0997 &   0.1689 &      0.1304 \\
+            6 &          0.0922 &   \textbf{0.1889} &      \textbf{0.1358} \\
+            8 &          \textbf{0.1001} &   0.1865 &      0.1327 \\
+\bottomrule
+\end{tabularx}
+\end{subtable}
+\hspace{\fill}
+\begin{subtable}{\columnwidth}
+\caption{Factorization models}
+\begin{tabularx}{\columnwidth}{rrrr}
+\toprule
+   Components &   Movielens 10M &   Amazon &   Goodbooks \\
+\midrule
+            2 &          0.0931 &   0.1343 &      0.0797 \\
+            4 &          \textbf{0.1026} &   0.1583 &      0.0840 \\
+            6 &          0.0864 &   0.1612 &      0.0811 \\
+            8 &          0.0879 &   \textbf{0.1696} &      \textbf{0.0853} \\
+\bottomrule
+\end{tabularx}
+\end{subtable}\end{table}
 We summarize the effect of increasing the number of taste mixture components in Table~\ref{tab:nummixtures}.

### 2017-12-13 paper/paper.tex
Make dot product notation more obvious.

@@
 \end{equation}
-where $b_j$ is the per-item bias term, $\sigma$ the elementwise sigmoid function, $\sigma\left(A_ie_j\right)$ gives the mixture probabilities, and $U_ie_j$ the recommendation scores given by each mixture component.
+where $b_j$ is the per-item bias term, $\sigma$ the elementwise sigmoid function, $\sigma\left(A_ie_j\right)$ gives the mixture probabilities, $U_ie_j$ the recommendation scores given by each mixture component, and $\cdot$ denotes the dot product.
 We assume identity variance matrices for all mixture components.

### 2018-01-18 paper/paper.tex
Correct sigmoid -> softmax. Change eqs (3) and (4) to post-multiply to yield row vectors.

--- 2017-12-13 paper/paper.tex
@@
 \end{equation}
-where $b_j$ is the per-item bias term, $\sigma$ the elementwise sigmoid function, $\sigma\left(A_ie_j\right)$ gives the mixture probabilities, and $U_ie_j$ the recommendation scores given by each mixture component.
+where $b_j$ is the per-item bias term, $\sigma$ the softmax function, $\sigma\left(A_ie_j\right)$ gives the mixture probabilities, $U_ie_j$ the recommendation scores given by each mixture component, and $\cdot$ denotes the vector dot product.
 We assume identity variance matrices for all mixture components.
@@
 \end{equation}
-Given $z_{it}$, the $m$-th row of $U_{it}$ and $A_{it}$ are given by
+Given the $1 \times k$ dimensional hidden state $z_{it}$, the $m$-th row of $U_{it}$ and $A_{it}$ are given by
 \begin{equation}
 \begin{aligned}
-  u^m_{it} &= W^U_mz_{it} + B^U_m\\
-  a^m_{it} &= W^A_mz_{it} + B^A_m,\\
+  u^m_{it} &= z_{it}W^U_m + B^U_m\\
+  a^m_{it} &= z_{it}W^A_m + B^A_m,\\
 \end{aligned}
@@
 \begin{aligned}
-  u^m_{i} &= W^U_mz_{i} + B^U_m\\
-  a^m_{i} &= W^A_mz_{i} + B^A_m.\\
+  u^m_{i} &= z_{i}W^U_m + B^U_m\\
+  a^m_{i} &= z_{i}W^A_m + B^A_m.\\
 \end{aligned}

### 2018-01-18 paper/paper.tex
Start reformulatinnng model section.

@@
 
+Within the classical matrix factorization framework, we experiment with two approaches to modelling taste mixtures.
+The first is the Embedding-Mixture Factorization (\textbf{EM-F}) model.
+In this model, a user is represented by two sets of $m$ $k$-dimensional latent vectors: one where the vectors correspond to their tastes, and one that describes how competent each taste is at describing a given item.
+When evaluating an item, the dot products of the second set of vectors with the item latent vector give the relative weights of each tastes from the first set that should be applied when evaluating the user's preference.
+Compared to a traditional factorization model, the EM-F model requires $2m$ as many parameters, giving it more expressive power at the expense of using more parameters.
+
+The second factorization model we test is the Projection-Mixture Factorization (\textbf{PM-F}) model.
+It extends the basic matrix factorization model by using linear projection layers on top of user latent vectors to model multiple tastes: instead of embedding $m$ taste vectors and $m$ mixture vectors directly, we use $2m$ $k \times k$ projection matrices to project a single $k$-dimensional user item vector into $m$ taste and $m$ mixture vectors.
+These matrices are shared between all users, making the model substantially more parsimonious.
+
 Our sequence-based model is the Mixture LSTM (\textbf{M-LSTM}) model.
@@
 By using a recurrent architecture, we capture information from the identity of the items the user interacted with as well as the order in which those interactions occurred.
-
-Within the classical matrix factorization framework, we experiment with two approaches to modelling taste mixtures.
-The first is the Projection-Mixture Factorization (\textbf{PM-F}) model.
-It extends the basic matrix factorization model by using linear projection layers on top of user latent vectors to model multiple tastes.
-In contrast, the Embedding-Mixture Factorization (\textbf{EM-F}) model directly embeds all the mixture parameters for each user.
-This makes it less parsimonious, but potentially more expressive.
 

### 2018-01-20 paper/paper.tex
Layour corrections, spelling.

@@
 \documentclass[sigconf]{acmart}
+\makeatletter
+\renewcommand\@formatdoi[1]{\ignorespaces}
+\makeatother
 
@@
 \usepackage{natbib}
+\usepackage{balance}
 
@@
 \date{\today}
-\settopmatter{printacmref=false, printccs=false, printfolios=true}
-\setcopyright{none}
-\acmConference[]{}{}{}
+\settopmatter{printacmref=true, printccs=false, printfolios=true}
+\setcopyright{rightsretained}
+\acmConference[Learn-IR]{Workshop on Learning From User Interactions, WSDM 2018}{February 5--9, 2018}{Los Angeles, United States}
+\keywords{Recommender Systems}
 
@@
 \begin{abstract}
-Most existing recommendation approaches implicitly treat user tastes as unimodal, resulting in an average-of-tastes representations when multiple distinct interests are present.
+Most existing recommendation approaches implicitly treat user tastes as unimodal, resulting in an average-of-tastes representation when multiple distinct interests are present.
 We show that appropriately modelling the multi-faceted nature of user tastes through a mixture-of-tastes model leads to large increases in recommendation quality.
 Our result holds both for deep sequence-based and traditional factorization models, and is robust to careful selection and tuning of baseline models.
-In sequence-based models, this improvement is achieved at a very modest cost in model complexity, making mixture-of-tastes models a straightforward improvement on existing baselines.
+In sequence-based models, this improvement is achieved at a very modest cost in model complexity, making mixture-of-tastes models a straightforward improvement upon existing baselines.
 \end{abstract}
@@
 Latent vector-based recommender models commonly represent users with a single latent vector per user \citep{koren2009bellkor, hu2008collaborative}.
-This representation, while simple and efficient to compute, neglects the fact that users have diverse sub-tastes, and that observed user interactions can be seen as manifestations of several distinct tastes or intents.
+This representation, while simple and efficient to compute, neglects the fact that users have diverse tastes, and that observed user interactions can be seen as manifestations of several distinct tastes or intents.
 These may either be stable facets of the user's preference (liking both horror and documentary movies, or both bluegrass and classical music), context-driven changes (preference for short-form TV content during the week but long-form cinematography during the weekend), or manifestations of phenomena like account sharing, where two different users, with correspondingly different tastes, share the same user account.
@@
 The second factorization model we test is the Projection-Mixture Factorization (\textbf{PM-F}) model.
-It extends the basic matrix factorization model by using linear projection layers on top of user latent vectors to model multiple tastes: instead of embedding $m$ taste vectors and $m$ mixture vectors directly, we use $2m$ $k \times k$ projection matrices to project a single $k$-dimensional user item vector into $m$ taste and $m$ mixture vectors.
+It extends the basic matrix factorization model by using linear projection layers in ocnjunction with latent vectors to model multiple tastes: instead of embedding $m$ taste vectors and $m$ mixture vectors for each user directly, we use $2m$ $k \times k$ projection matrices to project a single $k$-dimensional user item vector into $m$ taste and $m$ mixture vectors.
 These matrices are shared between all users, making the model substantially more parsimonious.
@@
 
-Formally, let $U_i$ be a $m \times k$ matrix representing the $m$ tastes of user $i$, and $A_i$ be a $m \times k$ matrix representing the affinities of each taste for representing particular items.
-The recommendation score for item $j$, represented by a $k \times 1$-dimensional embedding vector $e_j$, is then given by
+Formally, let $U_i$ be a $m \times k$ matrix representing the $m$ tastes of user $i$, and $A_i$ be a $m \times k$ matrix representing the affinities of each taste from $U_i$ for representing particular items.
+The recommendation score for item $j$, whose representation is given by a $k \times 1$-dimensional embedding vector $e_j$, is then given by
 \begin{equation}
@@
 \end{equation}
-where $W^U_m$, $W^A_M$ are the learned projection matrices, and $B^U_m$ and $B^A_m$ are bias terms.
+where $W^U_m$, $W^A_M$ are the learned projection matrices, and $B^U_m$ and $B^A_m$ contain bias terms.
 Both the projection matrices and the biases are common across all users, representing a modest increase in the total number of model parameters.
@@
 \end{equation}
-This keeps the number of model parameters small at a potential cost to model capacity.
+This keeps the number of model parameters small at a potential cost to model expressiveness.
 In contrast, the EM-F model embeds $U_i$ and $A_i$ directly.
@@
 Secondly, we wish to understand the extent to which the mixture-of-interests models are \emph{fragile}, in the sense of being highly sensitive to hyperparameter choices.
-Such fragile algorithms are potentially of lesser utility in industry applications, where the engineering cost of tuning and maintaing them may outweigh the accuracy benefits they bring.
+Such fragile algorithms are potentially of lesser utility in industry applications, where the engineering cost of tuning and maintaining them may outweigh the accuracy benefits they bring.
 
@@
 \begin{table}
-\caption{Mean reciprocal rank across all users/sequences in the test set.
+\caption{Mean reciprocal rank (MRR) across all users/sequences in the test set.
 Note that due to differences in experimental protocol, results between sequence-based and factorization models are not directly comparable.}
@@
 \subsection{Ranking quality}
-  \begin{figure*}[h]
+  \begin{figure*}[h!]
     \centering
@@
 In sequence-based models, our M-LSTM model outperforms the baseline LSTM model on all datasets.
-The performance gains are particularly large on the Amazon (26\%) datset and Goodbooks (17\%) datasets, and smaller, but still meaningful, on the Movielens 10M dataset (10\%).
+The performance gains are particularly large on the Amazon (26\%) dataset and Goodbooks (17\%) datasets, and smaller, but still meaningful, on the Movielens 10M dataset (10\%).
 This is consistent with our conjectures on the nature of the datasets.

### 2018-01-20 paper/paper.tex
Add Wang and Cho (2017) citation.

@@
 This allows much clearer separation of distinct meanings.
-This work is closest to the approach presented here.
 
@@
 However, the model is potentially disadvantaged by its inability to model strong \emph{dis}taste for a given class of items.
+
+The approach closest to the one presented here, but focusing solely on the sequence-based setting, is presented in \citet{wang2017attention}, and was published contemporaneously with the writing of this paper.
+The authors use pre-trained and fixed item embeddings, recurrent neural networks with attention to model user sequences, and either feedforward or recurrent decoders to generate taste mixture components.
+They find that both attention and recurrent decoders improve the recommendation results.
+We suggest that the two papers should be seen are mutually reinforcing and complementary, and suggest that a combination of both approaches (recurrent decoders and end-to-end embedding training) would be beneficial.
 

### 2018-01-29 paper/paper.tex

--- 2018-01-20 paper/paper.tex
@@
 
-\citet{athiwaratkun2017multimodal} extend this idea by representing words a mixtures of Gaussians.
+\citet{athiwaratkun2017multimodal} extend this idea by representing words as mixtures of Gaussians.
 This allows them to capture polysemy by modelling each word with several distinct, small-variance Gaussian distributions (rather than artificially inflating the variance of a single distribution).
 This allows much clearer separation of distinct meanings.
-This work is closest to the approach presented here.
 
@@
 However, the model is potentially disadvantaged by its inability to model strong \emph{dis}taste for a given class of items.
+
+The approach closest to the one presented here, but focusing solely on the sequence-based setting and using pre-trained item embeddings, is presented in \citet{wang2017attention}, and was published contemporaneously with the writing of this paper.
+The authors use recurrent neural networks with attention to model user sequences, and either feedforward or recurrent decoders to generate taste mixture components.
+Their experiments confirm that the addition of taste mixtures improves recommendation quality.
+We suggest that the two papers should be seen as mutually reinforcing and complementary, and that a combination of both approaches (recurrent decoders and end-to-end embedding training) would be beneficial.
 
@@
 The second factorization model we test is the Projection-Mixture Factorization (\textbf{PM-F}) model.
-It extends the basic matrix factorization model by using linear projection layers in ocnjunction with latent vectors to model multiple tastes: instead of embedding $m$ taste vectors and $m$ mixture vectors for each user directly, we use $2m$ $k \times k$ projection matrices to project a single $k$-dimensional user item vector into $m$ taste and $m$ mixture vectors.
+It extends the basic matrix factorization model by using linear projection layers in conjunction with latent vectors to model multiple tastes: instead of embedding $m$ taste vectors and $m$ mixture vectors for each user directly, we use $2m$ $k \times k$ projection matrices to project a single $k$-dimensional user item vector into $m$ taste and $m$ mixture vectors.
 These matrices are shared between all users, making the model substantially more parsimonious.

### 2018-02-14 mixture/results.py
Correctly calculate and display dataset popularity bias.
Fortunately, substantive results not changed.

@@ -278,7 +278,7 @@ def generate_dataset_table():
                           ('Amazon', get_amazon_dataset()),
                           ('Goodbooks', get_goodbooks_dataset())):
 
-        item_counts = dataset.tocoo().getnnz(axis=1)
+        item_counts = dataset.tocoo().getnnz(axis=0)
 
         print('Dataset {}, ratio: {:0,}'
               .format(name, np.percentile(item_counts, 95) / np.percentile(item_counts, 50)))
